### 2020-07-22 Model Statement/MSTOM Statement.tex
Redifinition of all JAGS models
Changed JAGS notation and added new scripts along with a PDF that explains these models.

\title{Multi-State Temporal Occupancy Modeling}

\date{\today}

\author{
  Brian Gerber \\
  University of Rhode Island
  \and
  Kimberly Rivera \\
  University of Rhode Island
 \and
  Mason Fidino \\
  Urban Wildlife Institute
}

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[round]{natbib}
\usepackage{wasysym}

\setlength{\parindent}{0cm}
\setlength{\parskip}{0em}

\begin{document}
\maketitle

\section{Introduction}
We are interested in understanding the spatial variability and drivers of temporal habitat-use of wild animal populations. Common approaches of studying habitat-use or species occurrence are exclusively focused on identifying the spatial drivers of where animals are and are not. Occupancy models are commonly used in these studies to leverage repeat independent surveys to account for sampling biases (e.g., the probability a species occurs or asymtotically uses a spatial unit of interest (site), but goes undetected; \citealt{occupancybook}). The when of where these animals use a site is commonly ignored; in other words, the diel activity of the animal is disregarded when investigating spatial variation in habitat-use. \\

Our interest is to understand the temporal structuring of habitat use, such as what are the spatial locations (and associated probabilities) that an animal uses during the day (sunrise to sunset), during the night (sunset to sunrise), or during the day and night (sunrise to sunrise). The exact temporal structure ultimately depends on the diel pattern of the species; for species that are highly diurnal, temporal categories of interest could be dawn, morning, afternoon, and dusk and for species that are highly noctural, temporal categories may be dusk, early night, late night, and dusk. Further, the why an animal uses a location during one time period or another is species-specific and depends on the life history strategy and behavior of the animal. This model could be useful to understand temporal habitat loss due to anthropogenic activity; for example, a species may use a site when considering the entire 24-hour diel period despite human influence, but change the time period of day they use the site (day only $\rightarrow$ night only) or use the site less by dropping activity during one part of the day (day and night $\rightarrow$ night only). \\

The utility of this modeling framework will be well suited for remote sampling of species of interest that is continuous throughout the 24-hour diel period, such as camera traps. However, this type of plotless sampling in a continuous environment presents some reinterpretation of parameters as animals move in and out a camera site. Specifically, occupancy probabilities do not reflect `instantaneous occurrence', but rather `asymptotic occurrence', or the probability a site is used at least once over the duration of sampling \citep{efford2012}. Further, the probability of detection at a camera site is a product of the probability a camera takes a picture of the species as it passes in front of the camera and the probability the species is available to be sampled. Most likely, the probability of availability is the main source of heterogeneity. As such, the parameter reflecting the probability of observing the species at a given site and survey occasion should likely \emph{a priori} considered to be different across temporal periods (e.g., day and night), unless the species is known to be active equivalently during these periods (\emph{}.g., cathemeral).\\

Below, we outline a series of multi-state temporal occupancy models (MSTOM) that can be used to understand temporal habitat-use. This model is a special form of the general multi-state occupancy model with state uncertainty \citep{MacKenzie2009}.

\section{Single Season MSTOM}
\subsection{Fully State Specific}
\subsubsection{Probability Model}
We assume each $i$ = 1...$N$ sites (spatial locations) are sampled with $j$ = 1....$K$ independent surveys. Each site is considered to be in one of $M$ mutually exclusive states during the survey period in which we can assume there are no changes in the state occurrence (i.e., closure). We will define four total states throughout ($M = 4$) where the states of site $i$ are 1) not occupied, 2) occupied during the day only, 3) occupied during night only, and 4) occupied during the night and day. State occurrence probabilities in the same order are,

%Perhaps not use the day, night, ND notation as it is confused with below. Do I need it?
\begin{equation}
\boldsymbol{\psi} = \begin{bmatrix} \psi^1 & \psi^2 & \psi^3 & \psi^4 \end{bmatrix}, %= \begin{bmatrix} \psi^0 & \psi^{\text{Day}} & \psi^{\text{Night}} & \psi^{\text{ND}} \end{bmatrix}
\end{equation}

where  $\psi^{1} = 1 - \psi^{2} - \psi^{3} - \psi^{4}$ and $\boldmath{1} \cdot \boldsymbol{\psi} = \text{1}$. The overall marginal (regardless of the state) occurrence of the species is $\psi^{\bullet} = \psi^{2} + \psi^{3} + \psi^{4}$ The true occupancy state for site $i$ is defined by the latent variable,

\begin{equation}
\textbf{z}_{i} \sim \text{Categorical}(\boldsymbol{\psi})
\end{equation}
and is a 1 x $M$ matrix. For each true state, there is an order to which states may be observed, but this is not strictly hierarchical, as commonly applied in multi-state occupancy models \citep{nichols2007}. Specifically, if a site is in state 3 (night only), it can not be observed in state 2 (day only). Let,  $p^{m,l}_{j}$ be the probability of observing the occupancy state $l$, given the true state is $m$ in survey $j$. To maintain the nomenclature of our intended inference and assuming no site- or survey-level variation in detection probabilities, we can define the $p^{2,2}$ as the probability of detection during the day ($p^{\text{Day}}$), $p^{3,3}$ as the probability of detection during the night ($p^{\text{Night}}$), and the detection probability in state 4 (in order) as $pND^{1}$, $pND^{2}$, $pND^{3}$, and $pND^{4}$. The detection matrix for survey $j$ is $M$ x $M$ with the observed (columns) and true state (row) as,

\begin{equation}
\boldsymbol{p}_{j} = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 
									1-p^{\text{Day}} & p^{\text{Day}} & 0 & 0 \\ 
									1-p^{\text{Night}} & 0 & p^{\text{Night}} & 0\\
  								      p^{\text{ND1}} & p^{\text{ND2}} & p^{\text{ND3}} & p^{\text{ND4}}
  								      \end{bmatrix}
\end{equation}
where $p^{1,4} = p^{\text{ND1}} = 1-pND^{2} - pND^{3} -pND^{4}$ and $\boldsymbol{p}_{j} \cdot \boldsymbol{1} = 1$. The observed state at site $i$ in survey $j$ is a 1 x $M$ categorical random variable as,

\begin{equation}
\textbf{y}_{ij} \sim \text{Categorical}(\textbf{z}_{i} \cdot \boldsymbol{p}_{j}).
\end{equation}

Diffuse prior distributions could be defined as,
\begin{center}
\begin{align*}
\boldsymbol{\psi} &\sim \text{Dirichlet}(1,1,1,1)\\
p^{2,2}= p^{\text{Day}} &\sim \text{Beta}(1,1)\\
p^{3,3} = p_{j}^{\text{Night}}  &\sim \text{Beta}(1,1)\\
\boldsymbol{p}^{4 \bullet} = \boldsymbol{p^{\text{ND}\bullet}} &\sim \text{Dirichlet}(1,1,1,1).\\
\end{align*}
\end{center}

Under this model, there is no shared relationship between the state occupancy probability 4 (night and day) and states 2 (day only) and 3 (night only). There is also no relationship shared between the detection probabilities of the same states. This model would suggest that the detection or availability of the species at state 4 is possibly different then the detection in other states. This might be because of the species' behavior at sites used throughout the diel period is different than at the other sites. This model is coded for JAGS in the file `jags.multistate.occ.full.R'.

\subsubsection{Multinomial-logit Model 1}
To model site-level occurrence according to a single continuous covariates ($\mathbf{x}$) that effects state-level probabilities differently, we could specify this as,
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi_{i}})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
\psi^1_{i} &=\frac{\phi^1_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^2_{i} &=\frac{\phi^2_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^3_{i} &=\frac{\phi^3_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^4_{i} &=\frac{\phi^4_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}.\\
\phi^1_{i} &= 1\\
\phi^2_{i} &= e^{\alpha_{1}+\alpha_{2}\times x_{i}}\\
\phi^3_{i} &= e^{\alpha_{3}+\alpha_{4}\times x_{i}}\\
\phi^4_{i} &= e^{\alpha_{5}+\alpha_{6}\times x_{i}}\\
\end{align*}
\end{center}

A diffuse prior could be $\alpha_{\bullet} \sim \text{Logistic}(0,1).$ This model is coded for JAGS in the file `jags.multistate.occ.full.site.covs.R'.

\subsubsection{Multinomial-logit Model 2}
We could generalize the modeling of state-specific occupancy parameters using separate design matrices ($\textbf{x}_{i}^{\text{Day}}, \textbf{x}_{i}^{\text{Night}}, \textbf{x}_{i}^{\text{ND}}$) that for each site $i$ are  1 x Q$_{m}$ (being the number of columns) and associated vectors of coefficients ($\boldsymbol{\alpha}^{\text{Day}}, \boldsymbol{\alpha}^{\text{Night}}, \boldsymbol{\alpha}^{\text{ND}}$) that are $Q_{m}$ x 1, and specified as,
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi_{i}})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
\psi^1_{i} &=\frac{\phi^1_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^2_{i} &=\frac{\phi^2_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^3_{i} &=\frac{\phi^3_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^4_{i} &=\frac{\phi^4_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}.\\
\phi^1_{i} &= 1\\
\phi^2_{i} &= e^{\textbf{x}_{i}^{\text{Day}}\boldsymbol{\alpha}^{\text{Day}}}\\
\phi^3_{i} &= e^{\textbf{x}_{i}^{\text{Night}}\boldsymbol{\alpha}^{\text{Night}}}\\
\phi^4_{i} &= e^{\textbf{x}_{i}^{\text{ND}}\boldsymbol{\alpha}^{\text{ND}}}\\
\end{align*}
\end{center}

A diffuse prior for all coefficients could be done using $\text{Logistic}(0,1).$ This model allows state-specific modeling of occupancy probabilities by any combination of covaraites when specificed using a design matrix, which is common in linear regression modeling. This model is coded for JAGS in the file `jags.multistate.occ.full.site.covs.R'.

\subsection{Reduced MSTOM}
\subsubsection{Probability Model}
We might consider a simpler structure for $\boldsymbol{\psi}$ or $\boldsymbol{p}_{j}$ because of ecological or data sparsity reasons. It might be that sites used during the day and night (state 4) are not unique, but simply random and independently determined by the probability any site is used during the day ($\psi^{\text{Day.M}}$) and night ($\psi^{\text{Night.M}}$). Rather than exclusive state probabilities, $\psi^{\text{Day.M}}$ is the probability of using any site during the day, regardless of use or not use at night (marginal probability); further, $\psi^{\text{Night}}$ is the probability of using any site at night, regardless of use during the day.  We can redefine our occupancy parameters as,
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi})\\
\boldsymbol{\psi} &= \begin{bmatrix} \psi^1 & \psi^2 & \psi^3 & \psi^4 \end{bmatrix}\\
\psi^1 &=(1-\psi^{\text{Day.M}})(1-\psi^{\text{Night.M}})\\
\psi^2 &=\psi^{\text{Day.M}}(1-\psi^{\text{Night.M}})\\
\psi^3 &=(1-\psi^{\text{Day.M}})\psi^{\text{Night.M}}\\
\psi^4 &=\psi^{\text{Day.M}}\psi^{\text{Night.M}}.\\
\end{align*}
\end{center}

Similarly, we can simplify the detection matrix using the probability of detection during the day  ($p^{\text{Day.M}}$) and night ($p^{\text{Night.M}}$) as,
\begin{equation}
\boldsymbol{p}_{j} = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 
									1-p^{\text{Day.M}} & p^{\text{Day.M}} & 0 & 0 \\ 
									1-p^{\text{Night.M}} & 0 & p^{\text{Night.M}} & 0\\
  								      (1-p^{\text{Day.M}})(1-p^{\text{Night.M}}) & p^{\text{Day.M}}(1-p^{\text{Night.M}}) & (1-p^{\text{Day.M}})p^{\text{Night.M}} & p^{\text{Day.M}}p^{\text{Night.M}}
  								      \end{bmatrix}
\end{equation}
, where $p^{1,4}$ can also be specified as equal to $1- p^{\text{Day.M}}(1-p^{\text{Night.M}}) - (1-p^{\text{Day.M}})p^{\text{Night.M}} - p^{\text{Day.M}}p^{\text{Night.M}}$. Unless the species of interest is completely cathemeral and uses the day and night times in equal proportion, it is unlikely that $p^{\text{Day.M}}$ and $p^{\text{Night.M}}$ should be equivalent. We could use the diffuse prior distributions of 
\begin{center}
\begin{align*}
\psi^{\text{Day.M}} &\sim \text{Beta}(1,1)\\
\psi^{\text{Night.M}} &\sim \text{Beta}(1,1)\\
p^{\text{Day.M}} &\sim \text{Beta}(1,1)\\
p^{\text{Night.M}} &\sim \text{Beta}(1,1).\\
\end{align*}
\end{center}
This model is coded for JAGS in the file `jags.multistate.occ.reduced.R'.

\subsection{Logit Model 1}
We could consider the effects of a site-level covariate (\textbf{x}) on the marginal occurrence probabilities $(\psi^{\text{Day.M}}, \psi^{\text{Night.M}})$ by specifying,

\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi}_{i})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}\\
\psi^1_{i} &=(1-\psi^{\text{Day.M}}_{i})(1-\psi^{\text{Night.M}}_{i})\\%1-\psi^2-\psi^3-\psi^4\\
\psi^2_{i} &=\psi^{\text{Day.M}}_{i}(1-\psi^{\text{Night.M}}_{i})\\
\psi^3_{i} &=(1-\psi^{\text{Day.M}}_{i})\psi^{\text{Night.M}}_{i}\\
\psi^4_{i} &=\psi^{\text{Day.M}}_{i}\psi^{\text{Night.M}}_{i}\\
\text{logit}(\psi^{\text{Day.M}}_{i}) &= \alpha_{1}^{\text{Day}}+\alpha_{2}^{\text{Day}}\times x_{i}\\
\text{logit}(\psi^{\text{Night.M}}_{i}) &= \alpha_{1}^{\text{Night}}+\alpha_{2}^{\text{Night}}\times x_{i}\\
\end{align*}
\end{center}

Diffuse prior distributions for all $\alpha$ parameters could be $\text{Logistic}(0,1)$. This model is coded for JAGS in the file `jags.multistate.occ.reduced.site.covs.R'

\subsubsection{Logit Model 2}
We could generalize the modeling of marginal occupancy parameters using separate design matrices ($\textbf{x}_{i}^{\text{Day}}, \textbf{x}_{i}^{\text{Night}}$) that for each site $i$ are  1 x Q$_{m}$ (being the number of columns) and associated vectors of coefficients ($\boldsymbol{\alpha}^{\text{Day}}, \boldsymbol{\alpha}^{\text{Night}}$) that are $Q_{m}$ x 1, and specified as,

\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi_{i}})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
\psi^1_{i} &=(1-\psi^{\text{Day.M}}_{i})(1-\psi^{\text{Night.M}}_{i})\\%1-\psi^2-\psi^3-\psi^4\\
\psi^2_{i} &=\psi^{\text{Day.M}}_{i}(1-\psi^{\text{Night.M}}_{i})\\
\psi^3_{i} &=(1-\psi^{\text{Day.M}}_{i})\psi^{\text{Night.M}}_{i}\\
\psi^4_{i} &=\psi^{\text{Day.M}}_{i}\psi^{\text{Night.M}}_{i}\\
\text{logit}(\psi^{\text{Day.M}}_{i}) &= \textbf{x}_{i}^{\text{Day}}\boldsymbol{\alpha}^{\text{Day}}\\
\text{logit}(\psi^{\text{Night.M}}_{i}) &= \textbf{x}_{i}^{\text{Night}}\boldsymbol{\alpha}^{\text{Night}}\\
\end{align*}
\end{center}
Diffuse prior distributions for all $\alpha$ parameters could be $\text{Logistic}(0,1)$. This model is coded for JAGS in the file `jags.multistate.occ.reduced.site.covs.by.state.R'.

\subsubsection{Multinomial-logit Mode 1}
An alternative reduced model specification is to use the multinomial-logit parameterization, which can constrain relationships among state probabilties and include covariates. This model could parameterized as, 
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi_{i}})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
\psi^1_{i} &=\frac{\phi^1_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^2_{i} &=\frac{\phi^2_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^3_{i} &=\frac{\phi^3_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
\psi^4_{i} &=\frac{\phi^4_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}.\\
\phi^1_{i} &= 1\\
\phi^2_{i} &= e^{\alpha_{1}}\\
\phi^3_{i} &= e^{\alpha_{2}}\\
\phi^4_{i} &= e^{\alpha_{1}+\alpha_{2}}\\
\end{align*}
\end{center}

The detection matrix could also be specified similarly on the logit scale as,

\begin{equation}
\boldsymbol{q}_{j} = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 
									1 & e^{\beta_{1}} & 0 & 0 \\ 
									1 & 0 & e^{\beta_{2}} & 0\\
  								      1 & e^{\beta_{1}} & e^{\beta_{2}} & e^{\beta_{1}+\beta_{2}}  
  								      \end{bmatrix}
\end{equation}
and where the detection parameters are derived as,
\begin{center}
\begin{align*}
p^{\text{Day}} &= \frac{e^{\beta_{1}}}{1+e^{\beta_{1}}+0+0}\\
p^{\text{Night}} &= \frac{e^{\beta_{2}}}{1+0+e^{\beta_{2}}+0}\\
 p^{\text{ND4}} &= \frac{e^{\beta_{1}+\beta{2}}}{1+e^{\beta_{1}}+e^{\beta_{2}}+e^{\beta_{1}+\beta{2}}}
\end{align*}
\end{center}

This model could be easily extended to include a difference of occupancy and detection in state 4 by a small modification of $\phi^4_{i} = e^{\alpha_{1}+\alpha_{2}+\alpha_{3}}$ and the matrix cell $q^{4,4}$ to $e^{\beta_{1}+\beta_{2}+\beta_{3}} $. A diffuse prior for all coefficients could be done using $\text{Logistic}(0,1).$ This model is coded for JAGS in the file `jags.multistate.occ.reduced.alt.R'.\\

\subsection{Null MSTOM}
\subsubsection{Probability Model}
Furthermore, we could consider a multi-state model in which there is no spatial variability in temporal occurrence. This could be used as a null model in a model comparison framework to assess the importance of this type of heterogeneity. The state-specific occurrence probabilities can be described by the overall or marginal (regardless of state) occurrence probability ($\psi^{\bullet}$); since we assume the states do not matter, we then assume the probability of occurrence in each state is the same. The same is true for the overall or marginal probability of detection ($p^{\bullet}$). This model would be essentially a single state (occurrence or not occurrence) model. We can do so by defining,

\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi})\\
\boldsymbol{\psi} &= \begin{bmatrix} \psi^1 & \psi^2 & \psi^3 & \psi^4 \end{bmatrix}.\\
\psi^1 &= 1-\psi^{\bullet}\\
\psi^2 &=\frac{\psi^{\bullet}}{3}\\
\psi^3 &=\frac{\psi^{\bullet}}{3}\\
\psi^4 &=\frac{\psi^{\bullet}}{3}\\
\boldsymbol{\psi} &= \begin{bmatrix} \psi^1 & \psi^2 & \psi^3 & \psi^4 \end{bmatrix}.
\end{align*}
\end{center}

Similarly, we can define the detection matrix using $p^{\bullet}$ as,
\begin{equation}
\boldsymbol{p}_{j} = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 
									1-p^{\bullet}_{j} & p^{\bullet}_{j} & 0 & 0 \\ 
									1-p^{\bullet}_{j} & 0 & p^{\bullet}_{j} & 0\\
  								      1-p^{\bullet}_{j} & \frac{p^{\bullet}_{j}}{3} & \frac{p^{\bullet}_{j}}{3} & \frac{p^{\bullet}_{j}}{3}
  								      \end{bmatrix}
\end{equation}
We could use the diffuse prior distributions of 
\begin{center}
\begin{align*}
\psi^{\bullet} &\sim \text{Beta}(1,1)\\
p^{\bullet}_{j} &\sim \text{Beta}(1,1).\\
\end{align*}
\end{center}

This model is coded for JAGS in the file `jags.multistate.occ.null.R'.

\subsubsection{Multinomial-logit Model 1}
An alternative specification uses the multinomial logit to transform occurrence and detection probabilities to the logit-scale while ensuring appropriate constrains on parameters (i.e., probabilities stay between 0 and 1). This model specification is more conducive to including covariates to model their effects on different parameters. To do so we can define occupancy parameters with a shared log-valued parameter ($\alpha$), such that 
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi})\\
\boldsymbol{\psi} &= \begin{bmatrix} \psi^1 & \psi^2 & \psi^3 & \psi^4 \end{bmatrix}\\
\psi^1 &=\frac{\phi^1}{\phi^1+ \phi^2+\phi^3+\phi^4}\\
\psi^2 &=\frac{\phi^2}{\phi^1+ \phi^2+\phi^3+\phi^4}\\
\psi^3 &=\frac{\phi^3}{\phi^1+ \phi^2+\phi^3+\phi^4}\\
\psi^4 &=\frac{\phi^4}{\phi^1+ \phi^2+\phi^3+\phi^4}.\\
\phi^1 &= 3\\
\phi^2 &= e^{\alpha}\\
\phi^3 &= e^{\alpha}\\
\phi^4 &= e^{\alpha}\\
\end{align*}
\end{center}

Similarly, we can define the detection matrix on the logit scale using a shared log-valued parameter ($\beta$), such that 
\begin{equation}
\boldsymbol{q}_{j} = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 
									1 &  e^{\beta} & 0 & 0 \\ 
									1 & 0 & e^{\beta} & 0\\
  								       3 & e^{\beta} & e^{\beta} & e^{\beta}
  								      \end{bmatrix}.
\end{equation}
The probability of detection can be derived as, 
\begin{equation}
p^{\bullet} =\frac{e^{\beta}}{1+e^{\beta}+0+0}
\end{equation}
We could use the diffuse prior distributions of 
\begin{center}
\begin{align*}
\alpha &\sim \text{Logistic}(0,1)\\
\beta &\sim \text{Logistic}(0,1).\\
\end{align*}
\end{center}

\subsubsection{Multinomial-logit Model 2}
We could consider the effects of a site-level covariate (\textbf{x}) on the occupancy probability by specifying,
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi}_{i})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}\\
\psi^1_{i} &=\frac{\phi^1}{\phi^1+ \phi^2+\phi^3+\phi^4}\\
\psi^2_{i} &=\frac{\phi^2}{\phi^1+ \phi^2+\phi^3+\phi^4}\\
\psi^3_{i} &=\frac{\phi^3}{\phi^1+ \phi^2+\phi^3+\phi^4}\\
\psi^4_{i} &=\frac{\phi^4}{\phi^1+ \phi^2+\phi^3+\phi^4}.\\
\phi^1_{i} &= 3\\
\phi^2_{i} &= e^{\alpha_{1}+\alpha_{2}\times x_{i}}\\
\phi^3_{i} &= e^{\alpha_{1}+\alpha_{2}\times x_{i}}\\
\phi^4_{i} &= e^{\alpha_{1}+\alpha_{2}\times x_{i}}\\
\end{align*}
\end{center}
Since we are ignoring the state designations, there is no reason to use different design matrices of each state-occupancy parameter nor a reason to estimate different parameters. This model is coded for JAGS in the file `jags.multistate.occ.null.site.covs.R'.

\subsubsection{Multinomial-logit Model 3}
We could generalize the modeling of the occupancy parameter using the same design matrix ($\textbf{x}_{i}$) that for each site $i$ that is 1 x Q (being the number of columns) and associated vectors of coefficients ($\boldsymbol{\alpha})$ that are $Q_{m}$ x 1, we specify
\begin{center}
\begin{align*}
\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi}_{i})\\
\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}\\
\psi^1_{i} &=\frac{\phi^1}{\phi^1+ \phi^2+\phi^3+\phi^4}\\
\psi^2_{i} &=\frac{\phi^2}{\phi^1+ \phi^2+\phi^3+\phi^4}\\
\psi^3_{i} &=\frac{\phi^3}{\phi^1+ \phi^2+\phi^3+\phi^4}\\
\psi^4_{i} &=\frac{\phi^4}{\phi^1+ \phi^2+\phi^3+\phi^4}.\\
\phi^1_{i} &= 3\\
\phi^2_{i} &= e^{\textbf{x}_{i}\boldsymbol{\alpha}}\\
\phi^3_{i} &= e^{\textbf{x}_{i}\boldsymbol{\alpha}}\\
\phi4_{i} &= e^{\textbf{x}_{i}\boldsymbol{\alpha}}.\\
\end{align*}
\end{center}

Since we are ignoring the state designations, there is no reason to use different design matrices of each state-occupancy parameter nor a reason to estimate different parameters. This model is coded for JAGS in the file `jags.multistate.occ.null.site.covs.by.state.R'.

\section{Model Selection}
To compare models (e.g., Full, Reduced, and Null MSTOM) we could use the conditional predictive ordinate (CPO; \citealt{hooten2015}). To do so, we need to define the integrated likelihood. First, lets define $I_{i}$ as an indicator for site $i$ that is a 1 when the logical argument is true and otherwise 0 (see below) and $\boldsymbol{f}_{i} =  \begin{bmatrix}f_{i}^1 f_{i}^2 f_{i}^3 f_{i}^4  \end{bmatrix}$ is matrix of the frequency of observations in each state across all $K$ surveys. Therefore,  assuming detection parameters are constant over each $j$ survey, the integrated likelihood is specified as,

\begin{center}
\begin{align*}
[\textbf{y}_{i}|\boldsymbol{\psi}_{i}, \boldsymbol{p}]  &= \left(\psi^1+\psi^2(1-p^{2,2})^{f_{i}^1}+ \psi^3(1-p^{3,3})^{f_{i}^1}+\psi^4(1-p^{1,4})^{f_{i}^1}\right)^{I_{\text{max}(\boldsymbol{y}_{i\cdot})\equiv 1}}\\
&\times  \left(\psi^2(p^{2,2})^{f_{i}^2}(1-p^{2,2})^{f_{i}^1}+\psi^4(p^{1,4})^{f_{i}^1}(p^{2,4})^{f_{i}^2}(p^{3,4})^{f_{i}^3}(p^{4,4})^{f_{i}^4}\right)^{I_{\text{max}(\boldsymbol{y}_{i\cdot})\equiv 2}} \\
&\times  \left(\psi^3(p^{3,3})^{f_{i}^3}(1-p^{3,3})^{f_{i}^1}+\psi^4(p^{1,4})^{f_{i}^1}(p^{2,4})^{f_{i}^2}(p^{3,4})^{f_{i}^3}(p^{4,4})^{f_{i}^4}\right)^{I_{\text{max}(\boldsymbol{y}_{i\cdot})\equiv 3 \wedge \forall \boldsymbol{y}_{i\cdot}\not\equiv 2}}\\
&\times  \left(\psi^4(p^{1,4})^{f_{i}^1}(p^{2,4})^{f_{i}^2}(p^{3,4})^{f_{i}^3}(p^{4,4})^{f_{i}^4}\right)^{I_{\text{max}(\boldsymbol{y}_{i\cdot})\equiv 4 \vee 2\in \boldsymbol{y}_{i\cdot}\wedge 3 \in \boldsymbol{y}_{i\cdot}}}
\end{align*}
\end{center}

Following \citealt{broms2016}, for each MCMC iteration $s$ = 1,..., $S$,
\begin{center}
\begin{align*}
\text{CPO}  &=-\sum_{i=1}^{N} \text{log}(CPO_{ij})\\
\text{CPO}_{ij}  &=\frac{S}{\sum_{s=1}^S [\textbf{y}_{i}|\boldsymbol{\psi}_{i}^{(s)}, \boldsymbol{p}^{(s)} ]^{-1}}\\
\end{align*}
\end{center}

\section{Goodness of fit}
We can evaluate the overall fit of the model using a Bayesian p-value comparing the observed ($D$) and predicted model deviance ($\tilde{D}$) as,
\begin{center}
\begin{align*}
%D^{(s)} &= -2\cdot\text{log}[\textbf{y}|\boldsymbol{\theta}^{(s)}]\\
D^{(s)} &= -2\cdot \sum_{i=1}^N \text{log}[\textbf{y}_{i}|\boldsymbol{\psi}_{i}^{(s)},\boldsymbol{p}^{(s)}]\\
\tilde{D}^{(s)} &= -2\cdot \sum_{i=1}^N \text{log}[\tilde{\textbf{y}}_{i}|\boldsymbol{\psi}_{i}^{(s)},\boldsymbol{p}^{(s)}]
\end{align*}
\end{center}
where the predicted occurrences and detections are sampled for each $s$ MCMC iteration as,
\begin{center}
\begin{align*}
\tilde{\boldsymbol{z}}^{(s)}_{i} &\sim \text{Categorical}(\boldsymbol{\psi}^{(s)}_{i})\\
\tilde{\boldsymbol{y}}^{(s)}_{ij} &\sim  \text{Categorical}(\tilde{\textbf{z}}_{i} \cdot \boldsymbol{p}_{j})\\
\end{align*}
\end{center}

The Bayesian P-value can be calculated as the proportion of times $D$ is greater than $\tilde{D}$. For a model that fits well, the p-value should be near 0.5; if the p-value is greater than 0.95 or less than 0.05 than the model foes not fit the data well.

\pagebreak

\begin{thebibliography}{100}
\bibitem[Broms et al., 2016]{broms2016}Broms, K. M., Hooten, M. B., \& Fitzpatrick, R. M. (2016). Model selection and assessment for multi‐species occupancy models. Ecology, 97(7), 1759-1770.

\bibitem[Efford and Dawson, 2012]{efford2012}Efford, M. G., \& Dawson, D. K. (2012). Occupancy in continuous habitat. Ecosphere, 3(4), 1-15.

\bibitem[Hooten and Hobbs, 2015]{hooten2015}Hooten, M. B., \& Hobbs, N. T. (2015). A guide to bayesian model selection for ecologists. Ecological Monographs, 85, 3–28.

\bibitem[Mackenzie et al., 2009]{MacKenzie2009}MacKenzie, D. I., Nichols, J. D., Seamans, M. E., \& Gutiérrez, R. J. (2009). Modeling species occurrence dynamics with multiple states and imperfect detection. Ecology, 90(3), 823-835.

\bibitem[Mackenzie et al., 2017]{occupancybook}MacKenzie, D. I., Nichols, J. D., Royle, J. A., Pollock, K. H., Bailey, L., \& Hines, J. E. (2018). Occupancy estimation and modeling: inferring patterns and dynamics of species occurrence. 2nd Edition. Elsevier.

\bibitem[Nichols et al., 2007]{nichols2007}Nichols, J. D., Hines, J. E., Mackenzie, D. I., Seamans, M. E., \& Gutierrez, R. J. (2007). Occupancy estimation and modeling with multiple states and state uncertainty. Ecology, 88(6), 1395-1400.
\end{thebibliography}

\end{document}

### 2020-07-22 AJB Fosa/AJB.MSTOM.modeling.script.r

@@ -70,10 +70,10 @@ ni <- 20000  ;       nt <- 2;    nb <- 4000;    nc <- 3;   adapt=4000
 ### Fit Model1 - Full model - No Covariates ############################# 
 
 #jags data input
-data.input <- list(y = y, R = dim(y)[1], T = dim(y)[2])
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2])
 
 # Parameters monitored
-params <- c("alpha","psi", "pNight", "pDay","pND","prob")
+params <- c("PSI", "pNight", "pDay","pND")
 
 #Fit the model to do adapt phase
 model.jags <- jags.model(file="JAGS/jags.multistate.occ.full.R", 
@@ -108,9 +108,9 @@ write.table(CPO.out,file="AJB Fosa/CPO.out.AJB.csv",append=TRUE,col.names = FALS
 
 ### Model2 -Reduced model - No Covariates ############################# 
 
-K=2 
-data.input <- list(y = y, R = dim(y)[1], T = dim(y)[2],K=K)
-params <- c("alpha", "pNight", "pDay","prob")  
+Q=2 
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2],Q = Q)
+params <- c("pNight", "pDay","PSI","psiNight","psiDay")  
 
 model.jags <- jags.model(file="JAGS/jags.multistate.occ.reduced.R", 
                          data = data.input,
@@ -136,16 +136,15 @@ gelman.diag(M2.reduced.no.covs,multivariate = FALSE)
 fit <- combine.mcmc(M2.reduced.no.covs)
 
 M2.red.no.covs.CPO=CPO.function(fit,y,"reduced")
-CPO.out=t(matrix(c("M2.red.no.covs.CPO",M2.red.no.covs.CPO)))
+CPO.out=t(matrix(c("M2.red.no.covs",M2.red.no.covs.CPO)))
 write.table(CPO.out,file="AJB Fosa/CPO.out.AJB.csv",append=TRUE,col.names = FALSE,sep=",",row.names = FALSE)
 
 ### Model3 -Null model - No Covariates ############################# 
 
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2])
+params <- c("alpha","beta", "p.overall", "PSI") 
 
-data.input <- list(y = y, R = dim(y)[1], T = dim(y)[2])
[…]

### 2020-07-22 AJB Fosa/CPO.function.r

@@ -9,7 +9,7 @@ CPO.function=function(fit,y,model.type){
   pNight=fit[,which(grepl("pNight",colnames(fit)))]
   
   if(model.type=="null"){
-    pDay=fit[,which(grepl("pdet",colnames(fit)))]
+    pDay=fit[,which(grepl("p.overall",colnames(fit)))]
     pNight=pDay
     
     mat=cbind(log(3),qlogis(pDay),qlogis(pDay),qlogis(pDay))
@@ -51,7 +51,7 @@ CPO.function=function(fit,y,model.type){
   for(k in 1:dim(lik.save)[1]){
     
     site.num=k
-    index.occ=which(apply(as.matrix(colnames(fit)), 1, FUN=function(x){grepl(paste("prob[",site.num,",",sep=""), x, fixed = TRUE)}))
+    index.occ=which(apply(as.matrix(colnames(fit)), 1, FUN=function(x){grepl(paste("PSI[",site.num,",",sep=""), x, fixed = TRUE)}))
     
     #For site q in order of states by 1,2,3,4
     

### 2020-07-22 JAGS/jags.multistate.occ.full.R

@@ -1,50 +1,50 @@
     model  { 
     # Priors
-    pNight ~ dunif(0, 1)  
-    pDay ~ dunif(0, 1)
-    for (i in 1:4) {
-      beta[i] ~ dgamma(1, 1)   # Induce Dirichlet prior
-      pND[i] <- beta[i]/sum(beta[])
+    pNight ~ dbeta(1, 1)  
+    pDay ~ dbeta(1, 1)
+    for (m in 1:4) {
+      beta[m] ~ dgamma(1, 1)   # Induce Dirichlet prior
+      pND[m] <- beta[m]/sum(beta[])
 
-      alpha[i] ~ dgamma(1, 1)   # Induce Dirichlet prior
-      psi[i] <- alpha[i]/sum(alpha[])
+      alpha[m] ~ dgamma(1, 1)   # Induce Dirichlet prior
+      psi[m] <- alpha[m]/sum(alpha[])
     }
     # Define state vector
-    for (s in 1:R){
-      prob[s,1] <- psi[1]
-      prob[s,2] <- psi[2]
-      prob[s,3] <- psi[3]
-      prob[s,4] <- psi[4]
+    for (i in 1:N){
+      PSI[i,1] <- psi[1]
+      PSI[i,2] <- psi[2]
+      PSI[i,3] <- psi[3]
+      PSI[i,4] <- psi[4]
     }
     # Define observation matrix
-    # Order of indices: true state, time, observed state
-    for (t in 1:T){
-    p[1,t,1] <- 1
-    p[1,t,2] <- 0
-    p[1,t,3] <- 0
-    p[1,t,4] <- 0
-    p[2,t,1] <- 1-pDay
-    p[2,t,2] <- pDay
-    p[2,t,3] <- 0
-    p[2,t,4] <- 0
-    p[3,t,1] <- 1-pNight
-    p[3,t,2] <- 0
-    p[3,t,3] <- pNight
-    p[3,t,4] <- 0
-    p[4,t,1] <- pND[1]
-    p[4,t,2] <- pND[2]
-    p[4,t,3] <- pND[3]
-    p[4,t,4] <- pND[4]
+    # Order of indices: true state, survey occ, observed state
+    for (j in 1:K){
+    p[1,j,1] <- 1
+    p[1,j,2] <- 0
+    p[1,j,3] <- 0
+    p[1,j,4] <- 0
+    p[2,j,1] <- 1-pDay
+    p[2,j,2] <- pDay
[…]

### 2020-07-22 JAGS/jags.multistate.occ.full.site.covs.R

@@ -1,57 +1,57 @@
     model  { 
     # Priors for day/night detection
-    pNight ~ dunif(0, 1)
-    pDay ~ dunif(0, 1)
+    pNight ~ dbeta(1, 1)
+    pDay ~ dbeta(1, 1)
     # Priors for day&night detection. Dirchlet prior forces pND for each 
     #state i to sum to 1.
-    for (i in 1:4) {
-      beta[i] ~ dgamma(1, 1)   # Induce Dirichlet prior
-      pND[i] <- beta[i]/sum(beta[])
+    for (m in 1:4) {
+      beta[m] ~ dgamma(1, 1)   # Induce Dirichlet prior
+      pND[m] <- beta[m]/sum(beta[])
     }
     #Define priors for alpha parameters on logit-scale
-    for (i in 1:K) {
-    alpha[i] ~ dlogis(0,1)
+    for (q in 1:Q) { #Q is the number of estimated parameters
+    alpha[q] ~ dlogis(0,1)
     }
     # Define state vector for each s site
-    for (s in 1:R){
-      phi[s,1] <- 1
-      phi[s,2] <- exp(alpha[1]+alpha[2]*cov1[s])
-      phi[s,3] <- exp(alpha[3]+alpha[4]*cov1[s])
-      phi[s,4] <- exp(alpha[5]+alpha[6]*cov1[s])
-      prob[s,1]     <- phi[s,1]/sum(phi[s,])
-      prob[s,2]     <- phi[s,2]/sum(phi[s,])
-      prob[s,3]     <- phi[s,3]/sum(phi[s,])
-      prob[s,4]     <- phi[s,4]/sum(phi[s,])
+    for (i in 1:N){
+      phi[i,1] <- 1
+      phi[i,2] <- exp(alpha[1]+alpha[2]*x[i])
+      phi[i,3] <- exp(alpha[3]+alpha[4]*x[i])
+      phi[i,4] <- exp(alpha[5]+alpha[6]*x[i])
+      PSI[i,1]     <- phi[i,1]/sum(phi[i,])
+      PSI[i,2]     <- phi[i,2]/sum(phi[i,])
+      PSI[i,3]     <- phi[i,3]/sum(phi[i,])
+      PSI[i,4]     <- phi[i,4]/sum(phi[i,])
     }
     # Define observation matrix
[…]

### 2020-07-22 JAGS/jags.multistate.occ.full.site.covs.by.state.R

@@ -1,63 +1,63 @@
     model  { 
     # Priors for day/night detection
-    pNight ~ dunif(0, 1)
-    pDay ~ dunif(0, 1)
+    pNight ~ dbeta(1, 1)
+    pDay ~ dbeta(1, 1)
     # Priors for day&night detection. 
     #state i to sum to 1.
-    for (i in 1:4) {
-      beta[i] ~ dgamma(1, 1)   # Induce Dirichlet prior
-      pND[i] <- beta[i]/sum(beta[])
+    for (m in 1:4) {
+      beta[m] ~ dgamma(1, 1)   # Induce Dirichlet prior
+      pND[m] <- beta[m]/sum(beta[])
     }
     #Define priors for alpha parameters on logit-scale
-    for (i in 1:K.day) {
-    alpha.day[i] ~ dlogis(0,1)
+    for (q in 1:Q.day) {
+    alpha.day[q] ~ dlogis(0,1)
     }
-    for (i in 1:K.night) {
-    alpha.night[i] ~ dlogis(0,1)
+    for (q in 1:Q.night) {
+    alpha.night[q] ~ dlogis(0,1)
     }
-    for (i in 1:K.nd) {
-    alpha.nd[i] ~ dlogis(0,1)
+    for (q in 1:Q.nd) {
+    alpha.nd[q] ~ dlogis(0,1)
     }
     # Define state vector for each s site
-    for (s in 1:R){
-      phi[s,1] <- 1
-      phi[s,2] <- exp(inprod(Xday[s,],alpha.day))
-      phi[s,3] <- exp(inprod(Xnight[s,],alpha.night))
-      phi[s,4] <- exp(inprod(Xnd[s,],alpha.nd))
-      prob[s,1] <- phi[s,1]/sum(phi[s,])
-      prob[s,2] <- phi[s,2]/sum(phi[s,])
-      prob[s,3] <- phi[s,3]/sum(phi[s,])
-      prob[s,4] <- phi[s,4]/sum(phi[s,])
+    for (i in 1:N){
+      phi[i,1] <- 1
+      phi[i,2] <- exp(inprod(Xday[i,],alpha.day))
+      phi[i,3] <- exp(inprod(Xnight[i,],alpha.night))
+      phi[i,4] <- exp(inprod(Xnd[i,],alpha.nd))
[…]

### 2020-07-22 JAGS/jags.multistate.occ.null.R

@@ -1,56 +1,45 @@
     model  { 
     # Priors
-    alpha ~ dlogis(0,1)
-    beta ~ dlogis(0,1)
+    psi.overall ~ dbeta(1,1)
+    p.overall ~ dbeta(1,1)
  #Define state vector for each s site
-    for (s in 1:R){
-      phi[s,1] <- 3
-      phi[s,2] <- exp(alpha)
-      phi[s,3] <- exp(alpha)
-      phi[s,4] <- exp(alpha)
-      #Save the site-level occupancy probabilties      
-      prob[s,1] <- phi[s,1]/sum(phi[s,])
-      prob[s,2] <- phi[s,2]/sum(phi[s,])
-      prob[s,3] <- phi[s,3]/sum(phi[s,])
-      prob[s,4] <- phi[s,4]/sum(phi[s,])
+    for (i in 1:N){
+      PSI[i,1] <- 1-psi.overall
+      PSI[i,2] <- psi.overall/3
+      PSI[i,3] <- psi.overall/3
+      PSI[i,4] <- psi.overall/3 
     }
     # Define observation matrix
     # Order of indices: true state, time, observed state
-    for (t in 1:T){
-    p[1,t,1] <- 1
-    p[1,t,2] <- 0
-    p[1,t,3] <- 0
-    p[1,t,4] <- 0
-    p[2,t,1] <- 1
-    p[2,t,2] <- exp(beta)
-    p[2,t,3] <- 0
-    p[2,t,4] <- 0
-    p[3,t,1] <- 1
-    p[3,t,2] <- 0
-    p[3,t,3] <- exp(beta)
-    p[3,t,4] <- 0
-    p[4,t,1] <- 3
-    p[4,t,2] <- exp(beta)
-    p[4,t,3] <- exp(beta)
-    p[4,t,4] <- exp(beta)
+    for (j in 1:K){
+    p[1,j,1] <- 1-p.overall
+    p[1,j,2] <- 0
+    p[1,j,3] <- 0
+    p[1,j,4] <- 0
+    p[2,j,1] <- 1-p.overall
+    p[2,j,2] <- p.overall
+    p[2,j,3] <- 0
+    p[2,j,4] <- 0
+    p[3,j,1] <- 1-p.overall
+    p[3,j,2] <- 0
+    p[3,j,3] <- p.overall
+    p[3,j,4] <- 0
+    p[4,j,2] <- p.overall/3
[…]

### 2020-07-22 JAGS/jags.multistate.occ.null.alt.R

@@ -1,0 +1,57 @@
+    model  { 
+    # Priors
+    alpha ~ dlogis(0,1)
+    beta ~ dlogis(0,1)
+ #Define state vector for each s site
+    for (i in 1:N){
+      phi[i,1] <- 3
+      phi[i,2] <- exp(alpha)
+      phi[i,3] <- exp(alpha)
+      phi[i,4] <- exp(alpha)
+      #Save the site-level occupancy probabilties      
+      PSI[i,1] <- phi[i,1]/sum(phi[i,])
+      PSI[i,2] <- phi[i,2]/sum(phi[i,])
+      PSI[i,3] <- phi[i,3]/sum(phi[i,])
+      PSI[i,4] <- phi[i,4]/sum(phi[i,])
+    }
+    # Define observation matrix
+    # Order of indices: true state, time, observed state
+    for (j in 1:K){
+    q[1,j,1] <- 1
+    q[1,j,2] <- 0
+    q[1,j,3] <- 0
+    q[1,j,4] <- 0
+    q[2,j,1] <- 1
+    q[2,j,2] <- exp(beta)
+    q[2,j,3] <- 0
+    q[2,j,4] <- 0
+    q[3,j,1] <- 1
+    q[3,j,2] <- 0
+    q[3,j,3] <- exp(beta)
+    q[3,j,4] <- 0
+    q[4,j,1] <- 3
+    q[4,j,2] <- exp(beta)
+    q[4,j,3] <- exp(beta)
+    q[4,j,4] <- exp(beta)
+    }
+    # State-space likelihood
+    # State equation: model of true states (z)
+    for (i in 1:N){
+    z[i] ~ dcat(PSI[i,])
+    }
+    # Observation equation
+    for (i in 1:N){
+    for (j in 1:K){ 
+    y[i,j] ~ dcat(q[z[i],j,])
+    } #j
+    } #i
+   #The prob of occupancy derived for site 1 in state 2.   
+   psi<-phi[1,2]/sum(phi[1,])
+   #Overall occurence is the sum of states 2,3, and 4. Here, fixed for site 1
+   psi.overall <- PSI[1,2]+PSI[1,3]+PSI[1,4]
+   #The probability of detection is equal to the sum of detection in state 2,3, and 4.
+   #These estimates come from the beta parameter. Here, I have fixed this to be occasion 1 (t=1).
[…]

### 2020-07-22 JAGS/jags.multistate.occ.null.site.covs.R

@@ -1,57 +1,57 @@
 model  { 
   # Priors
-  for(i in 1:K){
-    alpha[i]~dlogis(0,1)
+  for(q in 1:Q){
+    alpha[q]~dlogis(0,1)
   }
   beta ~ dlogis(0,1)
   #Define state vector for each s site
-  for (s in 1:R){
-    phi[s,1] <- 3
-    phi[s,2] <- exp(alpha[1]+alpha[2]*cov1[s])
-    phi[s,3] <- exp(alpha[1]+alpha[2]*cov1[s])
-    phi[s,4] <- exp(alpha[1]+alpha[2]*cov1[s])
+  for (i in 1:N){
+    phi[i,1] <- 3
+    phi[i,2] <- exp(alpha[1]+alpha[2]*x[i])
+    phi[i,3] <- exp(alpha[1]+alpha[2]*x[i])
+    phi[i,4] <- exp(alpha[1]+alpha[2]*x[i])
     #Save the site-level occupancy probabilties      
-    prob[s,1] <- phi[s,1]/sum(phi[s,])
-    prob[s,2] <- phi[s,2]/sum(phi[s,])
-    prob[s,3] <- phi[s,3]/sum(phi[s,])
-    prob[s,4] <- phi[s,4]/sum(phi[s,])
+    PSI[i,1] <- phi[i,1]/sum(phi[i,])
+    PSI[i,2] <- phi[i,2]/sum(phi[i,])
+    PSI[i,3] <- phi[i,3]/sum(phi[i,])
+    PSI[i,4] <- phi[i,4]/sum(phi[i,])
   }
   # Define observation matrix
   # Order of indices: true state, time, observed state
-  for (t in 1:T){
-    p[1,t,1] <- 1
-    p[1,t,2] <- 0
-    p[1,t,3] <- 0
-    p[1,t,4] <- 0
-    p[2,t,1] <- 1
-    p[2,t,2] <- exp(beta)
-    p[2,t,3] <- 0
-    p[2,t,4] <- 0
-    p[3,t,1] <- 1
-    p[3,t,2] <- 0
-    p[3,t,3] <- exp(beta)
-    p[3,t,4] <- 0
-    p[4,t,1] <- 3
-    p[4,t,2] <- exp(beta)
-    p[4,t,3] <- exp(beta)
-    p[4,t,4] <- exp(beta)
+  for (j in 1:K){
+    q[1,j,1] <- 1
[…]

### 2020-07-22 JAGS/jags.multistate.occ.null.site.covs.by.state.R

@@ -1,58 +1,58 @@
     model  { 
     # Priors
-    for(i in 1:K){
-      alpha[i]~dlogis(0,1)
+    for(q in 1:Q){
+      alpha[q]~dlogis(0,1)
     }
     beta ~ dlogis(0,1)
  #Define state vector for each s site
-    for (s in 1:R){
-      phi[s,1] <- 3
-      phi[s,2] <- exp(inprod(X[s,],alpha))
-      phi[s,3] <- exp(inprod(X[s,],alpha))
-      phi[s,4] <- exp(inprod(X[s,],alpha))
+    for (i in 1:N){
+      phi[i,1] <- 3
+      phi[i,2] <- exp(inprod(X[i,],alpha))
+      phi[i,3] <- exp(inprod(X[i,],alpha))
+      phi[i,4] <- exp(inprod(X[i,],alpha))
       #Save the site-level occupancy probabilties      
-      prob[s,1]  <- phi[s,1]/sum(phi[s,])
-      prob[s,2]  <- phi[s,2]/sum(phi[s,])
-      prob[s,3]  <- phi[s,3]/sum(phi[s,])
-      prob[s,4]  <- phi[s,4]/sum(phi[s,])
+      PSI[i,1]  <- phi[i,1]/sum(phi[i,])
+      PSI[i,2]  <- phi[i,2]/sum(phi[i,])
+      PSI[i,3]  <- phi[i,3]/sum(phi[i,])
+      PSI[i,4]  <- phi[i,4]/sum(phi[i,])
     }
     # Define observation matrix
     # Order of indices: true state, time, observed state
-    for (t in 1:T){
-    p[1,t,1] <- 1
-    p[1,t,2] <- 0
-    p[1,t,3] <- 0
-    p[1,t,4] <- 0
-    p[2,t,1] <- 1
-    p[2,t,2] <- exp(beta)
-    p[2,t,3] <- 0
-    p[2,t,4] <- 0
-    p[3,t,1] <- 1
-    p[3,t,2] <- 0
-    p[3,t,3] <- exp(beta)
-    p[3,t,4] <- 0
-    p[4,t,1] <- 3
-    p[4,t,2] <- exp(beta)
-    p[4,t,3] <- exp(beta)
-    p[4,t,4] <- exp(beta)
+    for (j in 1:K){
+    q[1,j,1] <- 1
+    q[1,j,2] <- 0
+    q[1,j,3] <- 0
+    q[1,j,4] <- 0
+    q[2,j,1] <- 1
[…]

### 2020-07-22 JAGS/jags.multistate.occ.reduced.R

@@ -1,45 +1,45 @@
     model  { 
     # Priors
-    pNight ~ dunif(0, 1)
-    pDay ~ dunif(0, 1)
-    psiNight ~ dunif(0, 1)
-    psiDay ~ dunif(0, 1)
+    pNight ~ dbeta(1, 1)
+    pDay ~ dbeta(1, 1)
+    psiNight ~ dbeta(1, 1)
+    psiDay ~ dbeta(1, 1)
     # Define state vector
-    for (s in 1:R){
-    prob[s,1] <-  (1-psiDay)*(1-psiNight)                  #1-(psiDay-psiDay*psiNight)-(psiNight-psiDay*psiNight)-psiDay*psiNight 
-    prob[s,2] <-  psiDay*(1-psiNight)  
-    prob[s,3] <-  psiNight*(1-psiDay)
-    prob[s,4] <-  psiDay*psiNight
+    for (i in 1:N){
+    PSI[i,1] <-  (1-psiDay)*(1-psiNight)      #1-(psiDay-psiDay*psiNight)-(psiNight-psiDay*psiNight)-psiDay*psiNight 
+    PSI[i,2] <-  psiDay*(1-psiNight)  
+    PSI[i,3] <-  psiNight*(1-psiDay)
+    PSI[i,4] <-  psiDay*psiNight
     }
     # Define observation matrix
-    # Order of indices: true state, time, observed state
-    for (t in 1:T){
-    p[1,t,1] <- 1
-    p[1,t,2] <- 0
-    p[1,t,3] <- 0
-    p[1,t,4] <- 0
-    p[2,t,1] <- 1-pDay
-    p[2,t,2] <- pDay
-    p[2,t,3] <- 0
-    p[2,t,4] <- 0
-    p[3,t,1] <- 1-pNight
-    p[3,t,2] <- 0
-    p[3,t,3] <- pNight
-    p[3,t,4] <- 0
-    p[4,t,1] <- (1-pDay)*(1-pNight)
-    p[4,t,2] <- pDay*(1-pNight)
-    p[4,t,3] <- pNight*(1-pDay)
-    p[4,t,4] <- pDay*pNight 
+    # Order of indices: true state, survey occ, observed state
+    for (j in 1:K){
+    p[1,j,1] <- 1
+    p[1,j,2] <- 0
+    p[1,j,3] <- 0
+    p[1,j,4] <- 0
+    p[2,j,1] <- 1-pDay
[…]

### 2020-07-22 JAGS/jags.multistate.occ.reduced.site.covs.R

@@ -1,48 +1,48 @@
     model  { 
     # Priors
-    pNight ~ dunif(0, 1)
-    pDay ~ dunif(0, 1)
-    for(i in 1:K){
-      alpha[i]~dlogis(0,1)
+    pNight ~ dbeta(1, 1)
+    pDay ~ dbeta(1, 1)
+    for(q in 1:Q){
+      alpha[q]~dlogis(0,1)
     }
     # Define state vector
-    for (s in 1:R){
-    logit(aa[s]) <- alpha[1]+alpha[2]*cov1[s]
-    logit(bb[s]) <- alpha[3]+alpha[4]*cov1[s]
-    prob[s,2] <- aa[s]*(1-bb[s])
-    prob[s,3] <- bb[s]*(1-aa[s])
-    prob[s,4] <- aa[s]*bb[s]
-    prob[s,1] <- 1-prob[s,2]-prob[s,3]-prob[s,4]
+    for (i in 1:N){
+    logit(psiDay[i]) <- alpha[1]+alpha[2]*x[i]
+    logit(psiNight[i]) <- alpha[3]+alpha[4]*x[i]
+    PSI[i,1] <- (1-psiNight[i])*(1-psiDay[i])
+    PSI[i,2] <- psiDay[i]*(1-psiNight[i])
+    PSI[i,3] <- psiNight[i]*(1-psiDay[i])
+    PSI[i,4] <- psiDay[i]*psiNight[i]
     }
     # Define observation matrix
     # Order of indices: true state, time, observed state
-    for (t in 1:T){
-    p[1,t,1] <- 1
-    p[1,t,2] <- 0
-    p[1,t,3] <- 0
-    p[1,t,4] <- 0
-    p[2,t,1] <- 1-pDay
-    p[2,t,2] <- pDay
-    p[2,t,3] <- 0
-    p[2,t,4] <- 0
-    p[3,t,1] <- 1-pNight
-    p[3,t,2] <- 0
-    p[3,t,3] <- pNight
-    p[3,t,4] <- 0
-    p[4,t,1] <- (1-pNight)*(1-pDay)
-    p[4,t,2] <- pDay*(1-pNight)
-    p[4,t,3] <- pNight*(1-pDay)
-    p[4,t,4] <- pDay*pNight 
+    for (j in 1:K){
+    p[1,j,1] <- 1
+    p[1,j,2] <- 0
+    p[1,j,3] <- 0
+    p[1,j,4] <- 0
[…]

### 2020-07-22 JAGS/jags.multistate.occ.reduced.site.covs.by.state.R

@@ -1,53 +1,52 @@
     model  { 
     # Priors
-    pNight ~ dunif(0, 1)
-    pDay ~ dunif(0, 1)
+    pNight ~ dbeta(1, 1)
+    pDay ~ dbeta(1, 1)
     #Define priors for alpha parameters on logit-scale
-    for (i in 1:K.day) {
+    for (q in 1:Q.day) {
     alpha.day[i] ~ dlogis(0,1)
     }
-    for (i in 1:K.night) {
-    alpha.night[i] ~ dlogis(0,1)
+    for (q in 1:Q.night) {
+    alpha.night[q] ~ dlogis(0,1)
     }
     # Define state vector
-    for (s in 1:R){
-    logit(aa[s]) <-   inprod(Xday[s,],alpha.day)
-    logit(bb[s])   <- inprod(Xnight[s,],alpha.night)
-    prob[s,2] <- aa[s]*(1-bb[s])
-    prob[s,3] <- bb[s]*(1-aa[s])
-    prob[s,4] <- aa[s]*bb[s]
-    prob[s,1] <- 1-prob[s,2]-prob[s,3]-prob[s,4]
+    for (i in 1:N){
+    logit(psiDay[i]) <-   inprod(Xday[i,],alpha.day)
+    logit(psiNight[i])   <- inprod(Xnight[i,],alpha.night)
+    PSI[i,2] <- psiDay[i]*(1-psiNight[i])
+    PSI[i,3] <- psiNight[i]*(1-psiDay[i])
+    PSI[i,4] <- psiDay[i]*psiNight[i]
+    PSI[i,1] <- (1-psiDay[i])*(1-psiNight[i])  #1-PSI[i,2]-PSI[i,3]-PSI[i,4]
     }
     # Define observation matrix
     # Order of indices: true state, time, observed state
-    for (t in 1:T){
-    p[1,t,1] <- 1
-    p[1,t,2] <- 0
-    p[1,t,3] <- 0
-    p[1,t,4] <- 0
-    p[2,t,1] <- 1-pDay
-    p[2,t,2] <- pDay
-    p[2,t,3] <- 0
-    p[2,t,4] <- 0
-    p[3,t,1] <- 1-pNight
-    p[3,t,2] <- 0
-    p[3,t,3] <- pNight
-    p[3,t,4] <- 0
-    p[4,t,1] <- (1-pNight)*(1-pDay)
[…]

### 2020-07-22 RNP Fosa/CPO.function.r

@@ -1,4 +1,5 @@
-#Script to calculate CPO from a model
+#Script to calculate CPO from a model. This model can handle site level variation in state
+#occupancy parameters, but assumes no site or survey occasion variaion in detection
 
 CPO.function=function(fit,y,model.type){
   
@@ -9,7 +10,7 @@ CPO.function=function(fit,y,model.type){
   pNight=fit[,which(grepl("pNight",colnames(fit)))]
   
   if(model.type=="null"){
-    pDay=fit[,which(grepl("pdet",colnames(fit)))]
+    pDay=fit[,which(grepl("p.overall",colnames(fit)))]
     pNight=pDay
     
     mat=cbind(log(3),qlogis(pDay),qlogis(pDay),qlogis(pDay))
@@ -54,7 +55,7 @@ CPO.function=function(fit,y,model.type){
   for(k in 1:dim(lik.save)[1]){
     
     site.num=k
-    index.occ=which(apply(as.matrix(colnames(fit)), 1, FUN=function(x){grepl(paste("prob[",site.num,",",sep=""), x, fixed = TRUE)}))
+    index.occ=which(apply(as.matrix(colnames(fit)), 1, FUN=function(x){grepl(paste("PSI[",site.num,",",sep=""), x, fixed = TRUE)}))
     
     #For site q in order of states by 1,2,3,4
     

### 2020-07-22 RNP Fosa/RNP.MSTOM.modeling.script.r

@@ -75,13 +75,13 @@ hist(cov)
 cov1.unscaled=as.numeric(cov)
 cov1.scaled=as.numeric(scale(cov,center = TRUE,scale = TRUE))
 
-K=6 # number of alpha parameters
+Q=6 # number of alpha parameters
 
 #Package data
-data.input <- list(y = y, R = dim(y)[1], T = dim(y)[2],cov1=cov1.scaled,K=K)
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2],x=cov1.scaled,Q=Q)
 
 # Parameters monitored
-params <- c("alpha", "pNight", "pDay","pND","prob")
+params <- c("alpha", "pNight", "pDay","pND","PSI")
 
 #Fit the model to do adapt phase
 model.jags <- jags.model(file="JAGS/jags.multistate.occ.full.site.covs.R", 
@@ -131,13 +131,13 @@ hist(cov)
 cov1.unscaled=as.numeric(cov)
 cov1.scaled=as.numeric(scale(cov,center = TRUE,scale = TRUE))
 
-K=6 # number of alpha parameters
+Q=6 # number of alpha parameters
 
 #package Data
-data.input <- list(y = y, R = dim(y)[1], T = dim(y)[2], cov1=cov1.scaled,K=K)
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2], x=cov1.scaled,Q=Q)
 
 # Parameters monitored
-params <- c("alpha", "pNight", "pDay","pND","prob")  
+params <- c("alpha", "pNight", "pDay","pND","PSI")  
 
 
 #Fit the model to do adapt phase
@@ -182,12 +182,12 @@ hist(cov)
 cov1.unscaled=as.numeric(cov)
 cov1.scaled=as.numeric(scale(cov,center = TRUE,scale = TRUE))
 
-K=4 # number of alpha parameters
+Q=4 # number of alpha parameters
 
-data.input <- list(y = y, R = dim(y)[1], T = dim(y)[2],cov1=cov1.scaled,K=K)
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2],x=cov1.scaled,Q=Q)
 
 # Parameters monitored
-params <- c("alpha", "pNight", "pDay","prob") 
+params <- c("alpha", "pNight", "pDay","PSI") 
 
 model.jags <- jags.model(file="JAGS/jags.multistate.occ.reduced.site.covs.R", 
                          data = data.input,
[…]

### 2020-07-22 TEAM Fosa/TEAM.MSTOM.modeling.script.r

@@ -40,7 +40,7 @@ zst=rep(4,dim(y)[1])
 inits <- function(){list(z = zst)}
 
 # MCMC settings
-ni <- 8000  ;       nt <- 2;    nb <- 1000;    nc <- 3;   adapt=1000
+ni <- 20000  ;       nt <- 2;    nb <- 4000;    nc <- 3;   adapt=4000
 
 
 ### Fit Model1 - Full model - No Covariates ############################# 
@@ -80,7 +80,7 @@ fit <- combine.mcmc(M1.full.no.covs)
 
 M1.full.no.covs.CPO=CPO.function(fit,y,"full")
 CPO.out=t(matrix(c("M1.full.no.covs",M1.full.no.covs.CPO)))
-write.table(CPO.out,file="AJB Fosa/CPO.out.AJB.csv",append=TRUE,col.names = FALSE,sep=",",row.names = FALSE)
+write.table(CPO.out,file="TEAM Fosa/CPO.out.TEAM.csv",append=TRUE,col.names = FALSE,sep=",",row.names = FALSE)
 
 ### Model2 -Reduced model - No Covariates ############################# 
 
@@ -101,9 +101,9 @@ M2.reduced.no.covs <- coda.samples(model.jags, variable.names=params,
                                    n.iter=ni, 
                                    thin=nt,
                                    progress.bar="text")
-save(M2.reduced.no.covs,file="AJB Fosa/M2.reduced.no.covs.out")
+save(M2.reduced.no.covs,file="TEAM Fosa/M2.reduced.no.covs.out")
 
-#load("AJB Fosa/M2.reduced.no.covs")
+#load("TEAM Fosa/M2.reduced.no.covs")
 
 #plot(M2.reduced.no.covs,ask=TRUE)
 
@@ -113,11 +113,9 @@ fit <- combine.mcmc(M2.reduced.no.covs)
 
 M2.red.no.covs.CPO=CPO.function(fit,y,"reduced")
 CPO.out=t(matrix(c("M2.red.no.covs.CPO",M2.red.no.covs.CPO)))
-write.table(CPO.out,file="AJB Fosa/CPO.out.AJB.csv",append=TRUE,col.names = FALSE,sep=",",row.names = FALSE)
+write.table(CPO.out,file="TEAM Fosa/CPO.out.TEAM.csv",append=TRUE,col.names = FALSE,sep=",",row.names = FALSE)
 
[…]

### 2020-07-22 TEAM Fosa/TEAM.MSTOM.plotting.r

@@ -1,0 +1,86 @@
+#M1 is the most supported model
+#Load the model and plots results
+
+library(rjags)
+library(runjags)
+library(coda)
+library(bayesplot)
+library(ggplot2)
+
+
+load("TEAM Fosa/M1.full.no.covs.out")
+
+fit=combine.mcmc(M1.full.no.covs)
+
+#load the prepared data file
+#load("RNP Fosa/RNP.data")
+#covs=RNP.data[[2]]
+#cov=covs$DistTown
+#hist(cov)
+#cov1.unscaled=as.numeric(cov)
+#cov1.scaled=as.numeric(scale(cov,center = TRUE,scale = TRUE))
+
+
+#Plot detection parameters
+pDay=fit[,which(grepl("pDay",colnames(fit)))]
+pNight=fit[,which(grepl("pNight",colnames(fit)))]
+
+pND.0=fit[,which(grepl("pND",colnames(fit)))[1]]
+pND.D=fit[,which(grepl("pND",colnames(fit)))[2]]
+pND.N=fit[,which(grepl("pND",colnames(fit)))[3]]
+pND.ND=fit[,which(grepl("pND",colnames(fit)))[4]]
+
+det.matrix=cbind(pDay,
+                 pNight,
+                 pND.0,
+                 pND.D,
+                 pND.N,
+                 pND.ND)
+
+par(mfrow=c(1,2))
+plot(density(pDay),xlim=c(0,0.2),lwd=3)
+lines(density(pNight),xlim=c(0,0.2),lwd=3,col=2)
+plot(density(pND.D),xlim=c(0,0.2),lwd=3)
+lines(density(pND.N),xlim=c(0,0.2),lwd=3,col=2)
+lines(density(pND.ND),xlim=c(0,0.2),lwd=3,col=3)
+
+
+png(file="TEAM Fosa/RNP.fosa.det.parms.png",res=200,units = "in",height=8,width=8)
+color_scheme_set("red")
+mcmc_intervals(det.matrix, pars = colnames(det.matrix)[-3],xlab="Detection Parameters")+ labs(x = "Detection Probability",y="State")+
+     theme(text = element_text(size=20))
+dev.off()
+
+colnames(fit)
+#The first 6 columns are the alpha parameters
+fit.matrix=as.matrix(fit)
+
+png(file="TEAM Fosa/RNP.fosa.parms.png",res=200,units = "in",height=8,width=8)
+color_scheme_set("red")
+mcmc_areas(fit.matrix,
[…]

### 2020-07-22 simulation study/fit.sim.data.MSTOM.r

@@ -15,7 +15,7 @@
 #4: Night and Day
 
 
-#Below, we will fit three multi-state temporal occurence models, in the following order to the loaded 
+#Below, we will fit three multi-state temporal occurrence models, in the following order to the loaded 
 #simulated data: 
 #1) Null
 #2) Reduced
@@ -29,15 +29,13 @@
   library(jagsUI)
   library(rjags)
 
-#Set which data to load
-load.data="Null"
-
-if(load.data=="Full"){load("simulation study/sim.data.multistate.full") }
-if(load.data=="Red"){load("simulation study/sim.data.multistate.reduced") }
-if(load.data=="Null"){load("simulation study/sim.data.multistate.null") }
-
+  #Set which data to load "Full", "Red", or "Null
+  load.data="Red"
   
-  
+  if(load.data=="Full"){load("simulation study/sim.full.data") }
+  if(load.data=="Red"){load("simulation study/sim.reduced.data") }
+  if(load.data=="Null"){load("simulation study/sim.null.data") }
+
   # MCMC settings
   ni <- 10000  ;       nt <- 1;        nb <- 1000;  nc <- 1;  adapt <- 1000
 
@@ -48,7 +46,7 @@ if(load.data=="Null"){load("simulation study/sim.data.multistate.null") }
 ########################################
 ########################################
 #Start for loop
-for(q in 1:n.sim){
+for(q in 1:n.sim){ #
 
 #the first saved list object will be the data
   save.model=vector("list",2)
@@ -63,7 +61,7 @@ for(q in 1:n.sim){
 
 
 #Bundle data for jags
-  data.list <- list(y = obs.matrix, R = dim(obs.matrix)[1], T = dim(obs.matrix)[2])
+  data.list <- list(y = obs.matrix, N = dim(obs.matrix)[1], K = dim(obs.matrix)[2])
 
 #Initial values
   zst=rep(4,dim(data.list$y)[1])
@@ -71,9 +69,10 @@ for(q in 1:n.sim){
 
 ################################################################
 #Fit the Null Model 
-  params <- c("psi","pdet","alpha","beta","psi.overall")
+params <- c("psi","p.overall","alpha","beta","psi.overall")
+
 #Prepare the model and data
-  model.null <- jags.model(file="JAGS/jags.multistate.occ.null.R", 
+  model.null <- jags.model(file="JAGS/jags.multistate.occ.null.alt.R", 
                          data = data.list,
[…]

### 2020-07-22 simulation study/model.comparison.CPO.r

@@ -38,9 +38,9 @@ colnames(CPO.save)=c("Null","Reduced","Full")
 #data.reduced.3.models indicates that the data was simulated under the reduced model
 #data.full.3.models indicates that the data was simulated under the full model
 
-load(paste("simulation study/data.null.3.models",q,sep=""))
-#load(paste("simulation study/data.Red.3.models",q,sep=""))
-#load(paste("simulation study/data.full.3.models",q,sep=""))
+#load(paste("simulation study/fit.simdata.Null.",q,".out",sep=""))
+load(paste("simulation study/fit.simdata.Red.",q,".out",sep=""))
+#load(paste("simulation study/fit.simdata.Full.",q,".out",sep=""))
 
 #extract observed data
 obs.matrix=save.model$obs.matrix #observed data
@@ -73,7 +73,7 @@ obs.matrix=save.model$obs.matrix #observed data
     CPO.sites=n.mcmc/(apply(1/lik.save,1,sum))
     length(CPO.sites)
     CPO=(-1)*sum(log(CPO.sites))
-    CPO.save[q,m]=CPO
+    CPO.save[m]=CPO
 }    
     
 ##################################

### 2020-07-22 simulation study/null.model.comparison.r

@@ -26,13 +26,13 @@ library(jagsUI)
 library(rjags)
 
 #Load the simulated data - choose one
-load("simulation study/sim.data.multistate.null") 
+load("simulation study/sim.null.data") 
 
 #This is the simulated true logit coeficient
 alpha=1.2
 
 #This is the true probability of detection for the simulated data
-pdet.truth=0.7
+p.overall.truth=0.7
 
 sim.data$omega
 
@@ -52,8 +52,8 @@ ni <- 5000  ;       nt <- 1;        nb <- 1000;  nc <- 1;  adapt <- 1000
   #Bundle data for jags
   data.list <- list(
     y = obs.matrix, 
-    R = dim(obs.matrix)[1],
-    T = dim(obs.matrix)[2]
+    N = dim(obs.matrix)[1],
+    K = dim(obs.matrix)[2]
   )
   
   #Initial values
@@ -62,10 +62,10 @@ ni <- 5000  ;       nt <- 1;        nb <- 1000;  nc <- 1;  adapt <- 1000
   
   ################################################################
   #Fit the Null Model 
-  params <- c("psi","prob","pdet","alpha","beta","psi.overall")
+  params <- c("psi","PSI","p.overall","alpha","beta","psi.overall")
   #Prepare the model and data
   model.null <- jags.model(
-    file="JAGS/jags.multistate.occ.null.R", 
+    file="JAGS/jags.multistate.occ.null.alt.R", 
     data = data.list,
     inits=inits,
     n.chains = nc,
@@ -87,8 +87,8 @@ ni <- 5000  ;       nt <- 1;        nb <- 1000;  nc <- 1;  adapt <- 1000
   alpha.samples=model.null.fit$alpha[,,1]
   beta.samples=model.null.fit$beta[,,1]
   psi.samples=model.null.fit$psi[,,1]
-  psi.overall=model.null.fit$psi.overall[,,1]
-  pdet.samples=model.null.fit$pdet[,,1]
+  psi.overall.samples=model.null.fit$psi.overall[,,1]
+  p.overall.samples=model.null.fit$p.overall[,,1]
   
   #We need to derive state occupancy and detection probs
   psiDay=psi.samples
@@ -117,9 +117,9 @@ ni <- 5000  ;       nt <- 1;        nb <- 1000;  nc <- 1;  adapt <- 1000
   
   #add the mle estimate
   #plot the overall occupancy 
-  hist(psi.overall)
+  hist(psi.overall.samples)
[…]

### 2020-07-22 Model Statement/MSTOM Statement.tex
MSTOM pdf update

@@
 \begin{document}
+\setlength{\abovedisplayskip}{0pt}
+\setlength{\belowdisplayskip}{0pt}
+\setlength{\abovedisplayshortskip}{0pt}
+\setlength{\belowdisplayshortskip}{0pt}
 \maketitle
@@
 \psi^4_{i} &=\frac{\phi^4_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}.\\
+\end{align*}
+\end{center}
+
+\begin{center}
+\begin{align*}
 \phi^1_{i} &= 1\\
@@
 This model is coded for JAGS in the file `jags.multistate.occ.full.site.covs.R'.
+
+\subsubsection{Multinomial-logit Model 2}
+
+We could also model detection parameters by site (\textbf{x}) or survey specific (\textbf{V}) covarariates, such that 
+\begin{center}
+\begin{align*}
+\boldsymbol{q}_{ij} &= \begin{bmatrix} 1 & 0 & 0 & 0 \\ 
+									1 &  q^{2,2}_{ij} & 0 & 0 \\ 
+									1 & 0 & q^{3,3}_{ij} & 0\\
+  								       1 & q^{4,2}_{ij} & q^{4,3}_{ij} & q^{4,4}_{ij}
+  								      \end{bmatrix}\\
+q^{2,2}_{ij} &=e^{\beta^{\text{Day}}_{0} +\beta^{\text{Day}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
+q^{3,3}_{ij} &=e^{\beta^{\text{Night}}_{0} +\beta^{\text{Night}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
+q^{4,2}_{ij} &=e^{\beta^{\text{ND2}}_{0} +\beta^{\text{ND2}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
+q^{4,3}_{ij} &=e^{\beta^{\text{ND3}}_{0} +\beta^{\text{ND3}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
+q^{4,4}_{ij} &=e^{\beta^{\text{ND4}}_{0} +\beta^{\text{ND4}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}
+\end{align*}
+\end{center}
+and the probability of detection parameters can be derived as, 
+\begin{center}
+\begin{align*}
+p^{\text{Day}}_{ij} &=\frac{q^{2,2}_{ij}}{1+q^{2,2}_{ij}+0+0}\\
+p^{\text{Night}}_{ij} &=\frac{q^{3,3}_{ij}}{1+0+q^{3,3}_{ij}+0}\\
+p^{\text{ND1}}_{ij} &=\frac{1}{1+q^{4,2}_{ij}+q^{4,3}_{ij}+q^{4,4}_{ij}}\\
+p^{\text{ND2}}_{ij} &=\frac{q^{4,2}_{ij}}{1+q^{4,2}_{ij}+q^{4,3}_{ij}+q^{4,4}_{ij}}\\
+p^{\text{ND3}}_{ij} &=\frac{q^{4,3}_{ij}}{1+q^{4,2}_{ij}+q^{4,3}_{ij}+q^{4,4}_{ij}}\\
+p^{\text{ND4}}_{ij} &=\frac{q^{4,4}_{ij}}{1+q^{4,2}_{ij}+q^{4,3}_{ij}+q^{4,4}_{ij}}.
+\end{align*}
+\end{center}
+
+Diffuse priors for all $\beta$ parameters could be specifed as Logistic(0,1).
 

### 2020-07-22 JAGS/jags.multistate.occ.null.R
text edit
site s to site i in comment

@@ -2,7 +2,7 @@
     # Priors
     psi.overall ~ dbeta(1,1)
     p.overall ~ dbeta(1,1)
- #Define state vector for each s site
+ #Define state vector for each i site
     for (i in 1:N){
       PSI[i,1] <- 1-psi.overall
       PSI[i,2] <- psi.overall/3

### 2020-07-22 JAGS/jags.multistate.occ.null.alt.R

@@ -2,7 +2,7 @@
     # Priors
     alpha ~ dlogis(0,1)
     beta ~ dlogis(0,1)
- #Define state vector for each s site
+ #Define state vector for each i site
     for (i in 1:N){
       phi[i,1] <- 3
       phi[i,2] <- exp(alpha)

### 2020-07-22 JAGS/jags.multistate.occ.null.site.covs.R

@@ -4,7 +4,7 @@ model  {
     alpha[q]~dlogis(0,1)
   }
   beta ~ dlogis(0,1)
-  #Define state vector for each s site
+  #Define state vector for each i site
   for (i in 1:N){
     phi[i,1] <- 3
     phi[i,2] <- exp(alpha[1]+alpha[2]*x[i])

### 2020-07-22 JAGS/jags.multistate.occ.null.site.covs.by.state.R

@@ -4,7 +4,7 @@
       alpha[q]~dlogis(0,1)
     }
     beta ~ dlogis(0,1)
- #Define state vector for each s site
+ #Define state vector for each i site
     for (i in 1:N){
       phi[i,1] <- 3
       phi[i,2] <- exp(inprod(X[i,],alpha))

### 2020-07-22 TEAM Fosa/TEAM.MSTOM.modeling.script.r

@@ -46,10 +46,10 @@ ni <- 20000  ;       nt <- 2;    nb <- 4000;    nc <- 3;   adapt=4000
 ### Fit Model1 - Full model - No Covariates ############################# 
 
 #jags data input
-data.input <- list(y = y, R = dim(y)[1], T = dim(y)[2])
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2])
 
 # Parameters monitored
-params <- c("alpha","psi", "pNight", "pDay","pND","prob")
+params <- c("pNight", "pDay","pND","PSI")
 
 #Fit the model to do adapt phase
 model.jags <- jags.model(file="JAGS/jags.multistate.occ.full.R", 
@@ -85,8 +85,8 @@ write.table(CPO.out,file="TEAM Fosa/CPO.out.TEAM.csv",append=TRUE,col.names = FA
 ### Model2 -Reduced model - No Covariates ############################# 
 
 K=2 
-data.input <- list(y = y, R = dim(y)[1], T = dim(y)[2],K=K)
-params <- c("alpha", "pNight", "pDay","prob")  
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2],K=K)
+params <- c("pNight", "pDay","PSI")  
 
 model.jags <- jags.model(file="JAGS/jags.multistate.occ.reduced.R", 
                          data = data.input,
@@ -112,14 +112,14 @@ gelman.diag(M2.reduced.no.covs,multivariate = FALSE)
 fit <- combine.mcmc(M2.reduced.no.covs)
 
 M2.red.no.covs.CPO=CPO.function(fit,y,"reduced")
-CPO.out=t(matrix(c("M2.red.no.covs.CPO",M2.red.no.covs.CPO)))
+CPO.out=t(matrix(c("M2.red.no.covs",M2.red.no.covs.CPO)))
 write.table(CPO.out,file="TEAM Fosa/CPO.out.TEAM.csv",append=TRUE,col.names = FALSE,sep=",",row.names = FALSE)
 
 ### Model3 -Null model - No Covariates ############################# 
-data.input <- list(y = y, R = dim(y)[1], T = dim(y)[2])
-params <- c("alpha", "pdet", "prob") 
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2])
+params <- c("alpha", "p.overall", "PSI","psi.overall") 
 
[…]

### 2020-07-22 TEAM Fosa/TEAM.MSTOM.modeling.script.r
Update TEAM.MSTOM.modeling.script.r

@@ -84,8 +84,8 @@ write.table(CPO.out,file="TEAM Fosa/CPO.out.TEAM.csv",append=TRUE,col.names = FA
 
 ### Model2 -Reduced model - No Covariates ############################# 
 
-K=2 
-data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2],K=K)
+Q=2 
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2],Q=Q)
 params <- c("pNight", "pDay","PSI")  
 
 model.jags <- jags.model(file="JAGS/jags.multistate.occ.reduced.R", 
@@ -117,7 +117,7 @@ write.table(CPO.out,file="TEAM Fosa/CPO.out.TEAM.csv",append=TRUE,col.names = FA
 
 ### Model3 -Null model - No Covariates ############################# 
 data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2])
-params <- c("alpha", "p.overall", "PSI","psi.overall") 
+params <- c("alpha","beta", "p.overall", "PSI","psi.overall") 
 
 model.jags <- jags.model(file="JAGS/jags.multistate.occ.null.alt.R", 
                          data = data.input,
@@ -153,12 +153,12 @@ dim(survey.cov)
 
 #The X design matrix is the same for all states, but the effects are different
 Xday=Xnight=Xnd=survey.cov
-K.day=dim(Xday)[2] # number of alpha parameters for day state
-K.night=dim(Xnight)[2] # number of alpha parameters for night state
-K.nd=dim(Xnd)[2] # number of alpha parameters for night and day state
+Q.day=dim(Xday)[2] # number of alpha parameters for day state
+Q.night=dim(Xnight)[2] # number of alpha parameters for night state
+Q.nd=dim(Xnd)[2] # number of alpha parameters for night and day state
 
-data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2],K.day=K.day,
-                   K.night=K.night,K.nd=K.nd,Xday=Xday,Xnight=Xnight,Xnd=Xnd)
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2],Q.day=Q.day,
+                   Q.night=Q.night,Q.nd=Q.nd,Xday=Xday,Xnight=Xnight,Xnd=Xnd)
 
 params <- c("alpha.day","alpha.night","alpha.nd", "pNight", "pDay","pND","PSI")
 
[…]

### 2020-07-22 JAGS/jags.multistate.occ.reduced.site.covs.by.state.R
index fix on reduced model

@@ -4,7 +4,7 @@
     pDay ~ dbeta(1, 1)
     #Define priors for alpha parameters on logit-scale
     for (q in 1:Q.day) {
-    alpha.day[i] ~ dlogis(0,1)
+    alpha.day[q] ~ dlogis(0,1)
     }
     for (q in 1:Q.night) {
     alpha.night[q] ~ dlogis(0,1)

### 2020-07-24 AJB Fosa/AJB.MSTOM.modeling.script.r
minor updates

@@ -108,8 +108,7 @@ write.table(CPO.out,file="AJB Fosa/CPO.out.AJB.csv",append=TRUE,col.names = FALS
 
 ### Model2 -Reduced model - No Covariates ############################# 
 
-Q=2 
-data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2],Q = Q)
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2])
 params <- c("pNight", "pDay","PSI","psiNight","psiDay")  
 
 model.jags <- jags.model(file="JAGS/jags.multistate.occ.reduced.R", 
@@ -166,7 +165,7 @@ fit <- combine.mcmc(M3.null.no.covs)
 
 M3.null.no.covs.CPO=CPO.function(fit,y,"null")
 CPO.out=t(matrix(c("M3.null.no.covs",M3.null.no.covs.CPO)))
-write.table(CPO.out,file="CPO.out.AJB.csv",append=TRUE,col.names = FALSE,sep=",",row.names = FALSE)
+write.table(CPO.out,file="AJB Fosa/CPO.out.AJB.csv",append=TRUE,col.names = FALSE,sep=",",row.names = FALSE)
 
 ### Model1 -Full model - Survey Covariate ############################# 
 #Fit the full model with survey/year covariate (effect coding)
@@ -387,3 +386,61 @@ fit <- combine.mcmc(M3.null.covs.2)
 M3.null.covs.2.CPO=CPO.function(fit,y,"null")
 CPO.out=t(matrix(c("M3.null.covs.2",M3.null.covs.2.CPO)))
 write.table(CPO.out,file="AJB Fosa/CPO.out.AJB.csv",append=TRUE,col.names = FALSE,sep=",",row.names = FALSE)
+
+
+### Model1 -hybrid - no covariates - occupancy is null and detection is full############################# 
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2])
+
+params <- c("alpha","pND","psi.overall","pNight","pDay","PSI")
+model.jags <- jags.model(file="JAGS/jags.multistate.occ.null.det.null.R", 
+                         data = data.input,
+                         inits=inits,
+                         n.chains = nc,
+                         n.adapt=adapt)
+
+update(model.jags, n.iter=nb)
+
+M1.hybrid.no.covs <- coda.samples(model.jags, variable.names=params, 
+                              n.iter=ni, 
+                              thin=nt,
[…]

### 2020-07-24 AJB Fosa/AJB.MSTOM.plotting.r

@@ -3,8 +3,9 @@ library(runjags)
 library(bayesplot)
 library(ggplot2)
 rm(list=ls())
-load("AJB Fosa/M1.full.fit.plotting")
-
+#load("AJB Fosa/M1.full.fit.plotting")
+load("AJB Fosa/M1.full.no.covs.out")
+fit <- combine.mcmc(M1.full.no.covs)
 #Plot detection parameters
 pDay=fit[,which(grepl("pDay",colnames(fit)))]
 pNight=fit[,which(grepl("pNight",colnames(fit)))]
@@ -38,10 +39,10 @@ dev.off()
 
 head(fit)
 
-prob.state1=fit[,match("prob[1,1]",colnames(fit))]
-prob.state2=fit[,match("prob[1,2]",colnames(fit))]
-prob.state3=fit[,match("prob[1,3]",colnames(fit))]
-prob.state4=fit[,match("prob[1,4]",colnames(fit))]
+prob.state1=fit[,match("PSI[1,1]",colnames(fit))]
+prob.state2=fit[,match("PSI[1,2]",colnames(fit))]
+prob.state3=fit[,match("PSI[1,3]",colnames(fit))]
+prob.state4=fit[,match("PSI[1,4]",colnames(fit))]
 
 prob.matrix=cbind(prob.state1,prob.state2,prob.state3,prob.state4)
 hist(prob.state4)

### 2020-07-24 JAGS/jags.multistate.occ.null.det.null.R

@@ -1,0 +1,57 @@
+    model  { 
+    # Priors
+    alpha ~ dlogis(0,1)
+    pNight ~ dbeta(1, 1)  
+    pDay ~ dbeta(1, 1)
+    for (m in 1:4) {
+      beta[m] ~ dgamma(1, 1)   # Induce Dirichlet prior
+      pND[m] <- beta[m]/sum(beta[])
+    }
+    # Define state vector
+    for (i in 1:N){
+      phi[i,1] <- 3
+      phi[i,2] <- exp(alpha)
+      phi[i,3] <- exp(alpha)
+      phi[i,4] <- exp(alpha)
+      #Save the site-level occupancy probabilties      
+      PSI[i,1] <- phi[i,1]/sum(phi[i,])
+      PSI[i,2] <- phi[i,2]/sum(phi[i,])
+      PSI[i,3] <- phi[i,3]/sum(phi[i,])
+      PSI[i,4] <- phi[i,4]/sum(phi[i,])
+    }
+    # Define observation matrix
+    # Order of indices: true state, survey occ, observed state
+    for (j in 1:K){
+    p[1,j,1] <- 1
+    p[1,j,2] <- 0
+    p[1,j,3] <- 0
+    p[1,j,4] <- 0
+    p[2,j,1] <- 1-pDay
+    p[2,j,2] <- pDay
+    p[2,j,3] <- 0
+    p[2,j,4] <- 0
+    p[3,j,1] <- 1-pNight
+    p[3,j,2] <- 0
+    p[3,j,3] <- pNight
+    p[3,j,4] <- 0
+    p[4,j,1] <- pND[1]
+    p[4,j,2] <- pND[2]
+    p[4,j,3] <- pND[3]
+    p[4,j,4] <- pND[4]
+    }
+    # State-space likelihood
+    # State equation: model of true states (z)
+    for (i in 1:N){
+     z[i] ~ dcat(PSI[i,])
+    }
+    # Observation equation
+    for (i in 1:N){
+       for (j in 1:K){ 
+        y[i,j] ~ dcat(p[z[i],j,])
+       } #j
+    } #i
+   #The prob of occupancy derived for site 1 in state 2.   
+   psi<-phi[1,2]/sum(phi[1,])
+   #Overall occurence is the sum of states 2,3, and 4. Here, fixed for site 1
+   psi.overall <- PSI[1,2]+PSI[1,3]+PSI[1,4]
+}

### 2020-07-24 JAGS/jags.multistate.occ.null.site.covs.by.state.R

@@ -42,8 +42,8 @@
     z[i] ~ dcat(PSI[i,])
     }
     # Observation equation
-    for (s in 1:R){
-    for (t in 1:T){ 
+    for (i in 1:N){
+    for (j in 1:K){ 
     y[i,j] ~ dcat(q[z[i],j,])
     } #j
     } #i

### 2020-07-24 JAGS/jags.multistate.occ.red.det.null.R

@@ -1,0 +1,49 @@
+    model  { 
+    # Priors
+    psiNight ~ dbeta(1, 1)
+    psiDay ~ dbeta(1, 1)
+    pNight ~ dbeta(1, 1)  
+    pDay ~ dbeta(1, 1)
+    for (m in 1:4) {
+      beta[m] ~ dgamma(1, 1)   # Induce Dirichlet prior
+      pND[m] <- beta[m]/sum(beta[])
+    }
+    # Define state vector
+    for (i in 1:N){
+    PSI[i,1] <-  (1-psiDay)*(1-psiNight)      #1-(psiDay-psiDay*psiNight)-(psiNight-psiDay*psiNight)-psiDay*psiNight 
+    PSI[i,2] <-  psiDay*(1-psiNight)  
+    PSI[i,3] <-  psiNight*(1-psiDay)
+    PSI[i,4] <-  psiDay*psiNight
+    }
+    # Define observation matrix
+    # Order of indices: true state, survey occ, observed state
+    for (j in 1:K){
+    p[1,j,1] <- 1
+    p[1,j,2] <- 0
+    p[1,j,3] <- 0
+    p[1,j,4] <- 0
+    p[2,j,1] <- 1-pDay
+    p[2,j,2] <- pDay
+    p[2,j,3] <- 0
+    p[2,j,4] <- 0
+    p[3,j,1] <- 1-pNight
+    p[3,j,2] <- 0
+    p[3,j,3] <- pNight
+    p[3,j,4] <- 0
+    p[4,j,1] <- pND[1]
+    p[4,j,2] <- pND[2]
+    p[4,j,3] <- pND[3]
+    p[4,j,4] <- pND[4]
+    }
+    # State-space likelihood
+    # State equation: model of true states (z)
+    for (i in 1:N){
+     z[i] ~ dcat(PSI[i,])
+    }
+    # Observation equation
+    for (i in 1:N){
+       for (j in 1:K){ 
+        y[i,j] ~ dcat(p[z[i],j,])
+       } #j
+    } #i
+}

### 2020-07-24 TEAM Fosa/TEAM.MSTOM.modeling.script.r

@@ -84,8 +84,7 @@ write.table(CPO.out,file="TEAM Fosa/CPO.out.TEAM.csv",append=TRUE,col.names = FA
 
 ### Model2 -Reduced model - No Covariates ############################# 
 
-Q=2 
-data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2],Q=Q)
+data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2])
 params <- c("pNight", "pDay","PSI")  
 
 model.jags <- jags.model(file="JAGS/jags.multistate.occ.reduced.R", 
@@ -233,7 +232,7 @@ Q=6 # number of alpha parameters
 data.input <- list(y = y, N = dim(y)[1], K = dim(y)[2],Q=Q,X=X)
 
 # Parameters monitored
-params <- c("alpha","p.overall","psi.overall","psi", "PSI")
+params <- c("alpha","beta","p.overall","psi.overall","psi", "PSI")
 
 model.jags <- jags.model(file="JAGS/jags.multistate.occ.null.site.covs.by.state.R", 
                          data = data.input,

### 2020-07-24 TEAM Fosa/TEAM.MSTOM.plotting.r

@@ -55,10 +55,16 @@ colnames(fit)
 #The first 6 columns are the alpha parameters
 fit.matrix=as.matrix(fit)
 
+    index.occ1=which(apply(as.matrix(colnames(fit)), 1, FUN=function(x){grepl(paste("PSI[1,1]"), x, fixed = TRUE)}))
+    index.occ2=which(apply(as.matrix(colnames(fit)), 1, FUN=function(x){grepl(paste("PSI[1,2]"), x, fixed = TRUE)}))
+    index.occ3=which(apply(as.matrix(colnames(fit)), 1, FUN=function(x){grepl(paste("PSI[1,3]"), x, fixed = TRUE)}))
+    index.occ4=which(apply(as.matrix(colnames(fit)), 1, FUN=function(x){grepl(paste("PSI[1,4]"), x, fixed = TRUE)}))
+
+fit.psi=fit.matrix[,c(index.occ1,index.occ2,index.occ3,index.occ4)]
+    
 png(file="TEAM Fosa/RNP.fosa.parms.png",res=200,units = "in",height=8,width=8)
 color_scheme_set("red")
-mcmc_areas(fit.matrix,
-           pars = c("prob[1,1]", "prob[1,2]", "prob[1,3]", "prob[1,4]"),
+mcmc_areas(fit.psi,
            prob = 0.5,
            prob_outer=0.99) + geom_vline(xintercept=0, linetype="solid", 
                                          color = "black", size=1)+

### 2020-07-24 Diel_Occ.R
Add files via upload

@@ -1,0 +1,187 @@
+# Goals -----------------------------------------------------------------
+
+# Converting capture data (1's and 0's) into diel capture data (1,2,3,4)
+# 1 = 0 (not occupied), 2 = Only day, 3 = Only night, 4 = Day and Night use
+
+
+# soon to come ->function
+# df with columns Date, Time, Station 
+# occ_number <- function(df,lat=-15.206893, lon=49.603431, start, end, 
+#                        tz="Etc/GMT-3", occasion_step=6)
+
+# Set-up -----------------------------------------------------------------
+
+rm(list=ls())
+gc() #garbage collection
+
+library(tidyverse) # data manipulation 
+library(lubridate) # dates 
+library(suncalc) # sunlight and sunset 
+
+# R project location
+setwd("C:/Users/Kim Rivera/Documents/Summer Project/TOD data/Fosa")
+
+# load metadata 
+# Data Org = Date, Time (24 hr clock), Station (Camera) Name
+fosa=read_csv(file="AJB_2008_TOD.csv")
+fosa
+
+# Identify and delete/ edit incomplete data  
+
+# Organize Data Headers (if using read_csv)
+# colnames(fosa)=c("Date","Time","Station")
+
+
+# Organize dates ----------------------------------------------------------
+
+# ATTENTION: dates are in Eastern African Timezone 
+# if cameras malfunction and time is missing BUT
+# photos are available, use light to distinguish if photos are 'Night' or 'Day'
+# Here, raw data recorded 'Night' as PM and 'Day' as AM (see below)
+
+# mutate(date_transit = case_when(is.na(date_transit) & Time=="AM" ~ ymd_hms(Date,"12:00:00"),
+#                         is.na(date_transit) & Time=="PM" ~ ymd_hms(Date,"00:00:00"),
+#                         !is.na(date_transit) ~ date_transit))
+
+# combine date and time and set timezone
+fosa <- fosa %>% 
+  mutate(date = ymd_hms(paste(fosa$Date,fosa$Time), tz="Etc/GMT-3")) %>% 
+  arrange(date)
+
+
+# Transform time into 'Day' and 'Night' ---------------------------------------
+
[…]

### 2020-07-29 Model Statement/MSTOM Statement.tex
updated model statement file

@@
 \begin{document}
-\setlength{\abovedisplayskip}{0pt}
-\setlength{\belowdisplayskip}{0pt}
+\setlength{\abovedisplayskip}{-2pt}
+\setlength{\belowdisplayskip}{10pt}
 \setlength{\abovedisplayshortskip}{0pt}
-\setlength{\belowdisplayshortskip}{0pt}
+\setlength{\belowdisplayshortskip}{10pt}
 \maketitle
@@
   								      \end{bmatrix}\\
-q^{2,2}_{ij} &=e^{\beta^{\text{Day}}_{0} +\beta^{\text{Day}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
-q^{3,3}_{ij} &=e^{\beta^{\text{Night}}_{0} +\beta^{\text{Night}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
-q^{4,2}_{ij} &=e^{\beta^{\text{ND2}}_{0} +\beta^{\text{ND2}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
-q^{4,3}_{ij} &=e^{\beta^{\text{ND3}}_{0} +\beta^{\text{ND3}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
-q^{4,4}_{ij} &=e^{\beta^{\text{ND4}}_{0} +\beta^{\text{ND4}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}
+q^{2,2}_{ij} &=e^{\alpha^{\text{Day}}_{0} +\alpha^{\text{Day}}_{1}\times x_{i}+\alpha^{\text{Day}}_{2}\times v_{ij}}\\
+q^{3,3}_{ij} &=e^{\alpha^{\text{Night}}_{0} +\alpha^{\text{Night}}_{1}\times x_{i}+\alpha^{\text{Day}}_{2}\times v_{ij}}\\
+q^{4,2}_{ij} &=e^{\alpha^{\text{ND2}}_{0} +\alpha^{\text{ND2}}_{1}\times x_{i}+\alpha^{\text{Day}}_{2}\times v_{ij}}\\
+q^{4,3}_{ij} &=e^{\alpha^{\text{ND3}}_{0} +\alpha^{\text{ND3}}_{1}\times x_{i}+\alpha^{\text{Day}}_{2}\times v_{ij}}\\
+q^{4,4}_{ij} &=e^{\alpha^{\text{ND4}}_{0} +\alpha^{\text{ND4}}_{1}\times x_{i}+\alpha^{\text{Day}}_{2}\times v_{ij}}
 \end{align*}
@@
 
-Diffuse priors for all $\beta$ parameters could be specifed as Logistic(0,1).
+Diffuse priors for all $\alpha$ parameters could be specifed as Logistic(0,1).
 
@@
 Similarly, we can simplify the detection matrix using the probability of detection during the day  ($p^{\text{Day.M}}$) and night ($p^{\text{Night.M}}$) as,
+\vspace{10pt}
 \begin{equation}

### 2020-07-29 Data Processing/Diel_Occ_Func.R
Add files via upload

@@ -1,0 +1,135 @@
+# Function Purpose--------------------------------------------------------------
+# The diel_occ function allows single-state occupancy detection matrices to 
+# be converted into a 4 state matrix in order to better understand wildlife
+# diel activity. Rather than 0's indicating no detection and 1's indicating
+# an animal was detected, diel states are defined as the following:
+# 1 = not detected
+# 2 = detected during the day only (after sunrise, before sunset)
+# 3 = detected during the night only (after sunset, before sunrise)
+# 4 = detected during the day AND night 
+
+# Lets get started!
+
+# Clean Up ---------------------------------------------------------------------
+rm(list=ls())
+gc() #garbage collection
+
+# Required Libraries -----------------------------------------------------------
+library(tidyverse) # data manipulation 
+library(lubridate) # dates 
+library(suncalc) # sunlight and sunset 
+
+# Set your variables -----------------------------------------------------------
+time.zone = "Etc/GMT-3" # Time zone of your study site
+latitude = -15.206893 # Latitude of study site
+longitude = 49.603431 # Longitude of sudy site
+start.survey = "2008-09-02" # Starting date of survey
+end.survey = "2008-11-13" #Ending date of survey
+occasion.length = 6 # Number of days per occasion 
+station.ID = "AJB" # Name of Station 
+photo.data=read_csv(file="C:/Users/Kim Rivera/Documents/Summer Project/TOD data/AJB/Fosa/AJB_2008_TOD.csv") # CSV of 'Date', 'Time', and 'Station', of detections
+detection.matrix=read_csv(file="C:/Users/Kim Rivera/Documents/Summer Project/Capture Data/AJB/2008_occ.csv", na = ".") # Detection matrix
+
+# Run the function -------------------------------------------------------------
+diel_occ = function(photo.data, 
+                    detection.matrix, 
+                    time.zone, 
+                    latitude, 
+                    longitude, 
+                    start.survey, 
+                    end.survey, 
+                    station.ID,
+                    occasion.length){
+  # Organize dates -------------------------------------------------------------
+  # Here we combine and format the date-time in our defined time zone
+  photo.data <- photo.data %>% 
[…]

### 2020-07-30 JAGS/jags.dynamic.multistate.null.R
create dynamic null model
Following the naming nomenclature from the other models save for the transition matrix and the detection matrix.

@@ -1,0 +1,133 @@
+model{
+######
+# Detection model
+######
+for(i in 1:nsite) {
+  for(ti in 1:nyear) {
+    for(survey in 1:nsurvey) {
+      y[i, ti, survey] ~ dcat( rdm[i, ( 1:nout ) , z[i, ti], ti ] )
+    }
+  }
+}
+#######
+# Latent state model
+#######
+for(i in 1:nsite) {
+  # for first season
+  z[i, 1] ~ dcat( PSI[i, ( 1:nout )] )
+  for(t in 2:nyear) {
+    z[i, t] ~ dcat( tpm[i, ( 1:nout ) , z[ i, t-1], t-1] )
+  }
+}
+for( i in 1:nsite ) {
+######
+# Fill in all of the transition probabilities
+######
+######
+# Latent state for first season, PSI = transition vector for season 1
+######
+# First season probabilities for each state
+PSI[i, 1] <- 1 #------------------------------------------------------------|U
+PSI[i, 2] <- exp( psi[1, i] ) #---------------------------------------------|D
+PSI[i, 3] <- exp( psi[2, i] ) #---------------------------------------------|N
+PSI[i, 4] <- exp( psi[1, i] + psi[2, i] ) #---------------------------------|DN
+for(t in 2:nyear) {
+######
+# Latent state for dynamic part of model
+# tpm = transition probability matrix. All columns sum to 1.
+# dim(tpm)[1] = site i 
+# dim(tpm)[2] = state at time t
+# dim(tpm)[3] = state at time t-1
+# dim(tpm)[4] = time t (for temporally varying covariates)
+######
+# U to ...
+tpm[i, 1, 1, t-1] <- 1 #----------------------------------------------------|U
+tpm[i, 2, 1, t-1] <- exp( gam[1, i, t-1] ) #--------------------------------|D
+tpm[i, 3, 1, t-1] <- exp(                  gam[2, i, t-1] ) #---------------|N
+tpm[i, 4, 1, t-1] <- exp( gam[1, i, t-1] + gam[2, i, t-1] ) #---------------|DN
+# D to ...
+tpm[i, 1, 2, t-1] <- exp( eps[1, i, t-1] ) #--------------------------------|U
+tpm[i, 2, 2, t-1] <- 1 #----------------------------------------------------|D
[…]

### 2020-07-30 Model Statement/MSTOM Statement.tex
edits based on mason review
alter pdf

@@
 This model is a special form of the general multi-state occupancy model with state uncertainty \citep{MacKenzie2009}.
+Furthermore, the MSTOM Full model is equivalent to the original co-occurence occupancy model \citep{MacKenzie2004} with two-species and the equivalent four states.
+A demonstration of their equivalent estimation of the most probable parameters values (i.e., maximum likelihood estimates or mode of the posterior distribution) can be found in the R script, `MSTOM Full comparison with 2.species.occ.model.r'.
+Note, that while MacKenzie et al.
+2004 is known to have numerical optimization issues, especially when modeling covariates \citep{MacKenzie2004}, this issue does not persist when using a hierarchical Bayesian approach with markov chain monte carlo optimization.
 
@@
 \subsubsection{Multinomial-logit Model 1}
+Instead of focusing on estimating site-level occupancy probabilities, we could recast this model using a multinomial logit link as, 
+\begin{center}
+\begin{align*}
+\textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi})\\
+\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
+\psi^1 &=\frac{\phi^1_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
+\psi^2 &=\frac{\phi^2_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
+\psi^3 &=\frac{\phi^3_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
+\psi^4 &=\frac{\phi^4_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}.\\
+\end{align*}
+\end{center}
+
+\begin{center}
+\begin{align*}
+\phi^1 &= 1\\
+\phi^2 &= e^{\alpha_{1}}\\
+\phi^3 &= e^{\alpha_{2}}\\
+\phi^4 &= e^{\alpha_{1}+\alpha_{2}+\alpha_{3}}\\
+\end{align*}
+\end{center}
+
+A diffuse prior for all coefficients could be done using $\text{Logistic}(0,1).$ This model is coded for JAGS in the file `jags.multistate.occ.full.alt.R'.
+
+\subsubsection{Multinomial-logit Model 2}
 To model site-level occurrence according to a single continuous covariates ($\mathbf{x}$) that effects state-level probabilities differently, we could specify this as,
@@
 \phi^3_{i} &= e^{\alpha_{3}+\alpha_{4}\times x_{i}}\\
-\phi^4_{i} &= e^{\alpha_{5}+\alpha_{6}\times x_{i}}\\
+\phi^4_{i} &= e^{\alpha_{1}+\alpha_{2}\times x_{i}+\alpha_{3}+\alpha_{4}\times x_{i}+\alpha_{5}+\alpha_{6}\times x_{i}}\\
 \end{align*}
@@
 
-\subsubsection{Multinomial-logit Model 2}
+\subsubsection{Multinomial-logit Model 3}
 We could generalize the modeling of state-specific occupancy parameters using separate design matrices ($\textbf{x}_{i}^{\text{Day}}, \textbf{x}_{i}^{\text{Night}}, \textbf{x}_{i}^{\text{ND}}$) that for each site $i$ are  1 x Q$_{m}$ (being the number of columns) and associated vectors of coefficients ($\boldsymbol{\alpha}^{\text{Day}}, \boldsymbol{\alpha}^{\text{Night}}, \boldsymbol{\alpha}^{\text{ND}}$) that are $Q_{m}$ x 1, and specified as,
@@
 \phi^3_{i} &= e^{\textbf{x}_{i}^{\text{Night}}\boldsymbol{\alpha}^{\text{Night}}}\\
-\phi^4_{i} &= e^{\textbf{x}_{i}^{\text{ND}}\boldsymbol{\alpha}^{\text{ND}}}\\
+\phi^4_{i} &= e^{\textbf{x}_{i}^{\text{Day}}\boldsymbol{\alpha}^{\text{Day}}+\textbf{x}_{i}^{\text{Night}}\boldsymbol{\alpha}^{\text{Night}}+\textbf{x}_{i}^{\text{ND}}\boldsymbol{\alpha}^{\text{ND}}}\\
 \end{align*}
@@
 This model could be easily extended to include a difference of occupancy and detection in state 4 by a small modification of $\phi^4_{i} = e^{\alpha_{1}+\alpha_{2}+\alpha_{3}}$ and the matrix cell $q^{4,4}$ to $e^{\beta_{1}+\beta_{2}+\beta_{3}} $.
-A diffuse prior for all coefficients could be done using $\text{Logistic}(0,1).$ This model is coded for JAGS in the file `jags.multistate.occ.reduced.alt.R'.\\
+A diffuse prior for all coefficients could be done using $\text{Logistic}(0,1).$ \\
 
@@
 
+\bibitem[Mackenzie et al., 2004]{MacKenzie2004}MacKenzie, D.
+I., Bailey, L.
+L., \& Nichols, J.
+D.
+(2004).
+Investigating species co‐occurrence patterns when species are detected imperfectly.
+Journal of Animal Ecology, 73(3), 546-555.
+
 \bibitem[Mackenzie et al., 2009]{MacKenzie2009}MacKenzie, D.

### 2020-07-30 JAGS/jags.multistate.occ.full.site.covs.R

@@ -17,7 +17,9 @@
       phi[i,1] <- 1
       phi[i,2] <- exp(alpha[1]+alpha[2]*x[i])
       phi[i,3] <- exp(alpha[3]+alpha[4]*x[i])
-      phi[i,4] <- exp(alpha[5]+alpha[6]*x[i])
+      phi[i,4] <- exp(alpha[1]+alpha[2]*x[i]+
+                      alpha[3]+alpha[4]*x[i]+
+                      alpha[5]+alpha[6]*x[i])
       PSI[i,1]     <- phi[i,1]/sum(phi[i,])
       PSI[i,2]     <- phi[i,2]/sum(phi[i,])
       PSI[i,3]     <- phi[i,3]/sum(phi[i,])

### 2020-07-30 JAGS/jags.multistate.occ.full.site.covs.by.state.R

@@ -23,7 +23,7 @@
       phi[i,1] <- 1
       phi[i,2] <- exp(inprod(Xday[i,],alpha.day))
       phi[i,3] <- exp(inprod(Xnight[i,],alpha.night))
-      phi[i,4] <- exp(inprod(Xnd[i,],alpha.nd))
+      phi[i,4] <- exp(inprod(Xday[i,],alpha.day)+inprod(Xnight[i,],alpha.night)+inprod(Xnd[i,],alpha.nd))
       PSI[i,1] <- phi[i,1]/sum(phi[i,])
       PSI[i,2] <- phi[i,2]/sum(phi[i,])
       PSI[i,3] <- phi[i,3]/sum(phi[i,])

### 2020-07-30 JAGS/jags.multistate.occ.reduced.alt.R

@@ -1,24 +1,18 @@
     model  { 
     # Priors
-    beta1 ~ dlogis(0,1)
-    beta2 ~ dlogis(0,1)
-    beta3 ~ dlogis(0,1)
-    alpha1 ~ dlogis(0,1)
-    alpha2 ~ dlogis(0,1)
-    alpha3 ~ dlogis(0,1)
+    beta1 ~ dlogis(0, 1)
+    beta2 ~ dlogis(0, 1)
+    psiNight ~ dbeta(1, 1)
+    psiDay ~ dbeta(1, 1)
     # Define state vector
     for (i in 1:N){
-      phi[i,1] <- 1
-      phi[i,2] <- exp(alpha1)
-      phi[i,3] <- exp(alpha2)
-      phi[i,4] <- exp(alpha1+alpha2+alpha3)
-      PSI[i,1] <- phi[i,1]/phi[i,]
-      PSI[i,2] <- phi[i,2]/phi[i,]
-      PSI[i,3] <- phi[i,3]/phi[i,]
-      PSI[i,4] <- phi[i,4]/phi[i,]
+    PSI[i,1] <-  (1-psiDay)*(1-psiNight)                  #1-(psiDay-psiDay*psiNight)-(psiNight-psiDay*psiNight)-psiDay*psiNight 
+    PSI[i,2] <-  psiDay*(1-psiNight)  
+    PSI[i,3] <-  psiNight*(1-psiDay)
+    PSI[i,4] <-  psiDay*psiNight
     }
     # Define observation matrix
-    # Order of indices: true state, time, observed state
+    # Order of indices: true state, survey occ, observed state
     for (j in 1:K){
     q[1,j,1] <- 1
     q[1,j,2] <- 0
@@ -35,25 +29,20 @@
     q[4,j,1] <- 1
     q[4,j,2] <- exp(beta1)
     q[4,j,3] <- exp(beta2)
-    q[4,j,4] <- exp(beta1 + beta2 + beta3)
+    q[4,j,4] <- exp(beta1+beta2)
     }
     # State-space likelihood
     # State equation: model of true states (z)
     for (i in 1:N){
-     z[i] ~ dcat(phi[i,])
+    z[i] ~ dcat(PSI[i,])
     }
     # Observation equation
     for (i in 1:N){
-       for (j in 1:K){ 
-        y[i,j] ~ dcat(q[z[i],j,])
-       } #j
[…]

### 2020-07-30 JAGS/jags.multistate.occ.reduced.alt2.R

@@ -1,48 +1,0 @@
-    model  { 
-    # Priors
-    beta1 ~ dlogis(0, 1)
-    beta2 ~ dlogis(0, 1)
-    psiNight ~ dbeta(1, 1)
-    psiDay ~ dbeta(1, 1)
-    # Define state vector
-    for (i in 1:N){
-    PSI[i,1] <-  (1-psiDay)*(1-psiNight)                  #1-(psiDay-psiDay*psiNight)-(psiNight-psiDay*psiNight)-psiDay*psiNight 
-    PSI[i,2] <-  psiDay*(1-psiNight)  
-    PSI[i,3] <-  psiNight*(1-psiDay)
-    PSI[i,4] <-  psiDay*psiNight
-    }
-    # Define observation matrix
-    # Order of indices: true state, survey occ, observed state
-    for (j in 1:K){
-    q[1,j,1] <- 1
-    q[1,j,2] <- 0
-    q[1,j,3] <- 0
-    q[1,j,4] <- 0
-    q[2,j,1] <- 1
-    q[2,j,2] <- exp(beta1)
-    q[2,j,3] <- 0
-    q[2,j,4] <- 0
-    q[3,j,1] <- 1
-    q[3,j,2] <- 0
-    q[3,j,3] <- exp(beta2)
-    q[3,j,4] <- 0
-    q[4,j,1] <- 1
-    q[4,j,2] <- exp(beta1)
-    q[4,j,3] <- exp(beta2)
-    q[4,j,4] <- exp(beta1+beta2)
-    }
-    # State-space likelihood
-    # State equation: model of true states (z)
-    for (i in 1:N){
-    z[i] ~ dcat(PSI[i,])
-    }
-    # Observation equation
-    for (i in 1:N){
-    for (j in 1:K){ 
-    y[i,j] ~ dcat(q[z[i],j,])
-    } #j
-    } #i
-    pDay <- q[2,1,2]/sum(q[2,1,])
-    pNight <- q[3,1,3]/sum(q[3,1,])
-    pNDND <- q[4,1,4]/sum(q[4,1,])
-}

### 2020-07-30 simulation study/MSTOM Full comparison with 2.species.occ.model.r

@@ -1,0 +1,121 @@
+#Comparison of the origima; two species occupancy model (MacKenzie et al. 2004)
+#with that of the MSTOM Full model. These models are equivalent for 4 states (2-species).
+#However, the Bayesian MSTOM does not have the same estimation optimization issues, with
+#or without covariates.
+
+#MacKenzie, D. I., Bailey, L. L., & Nichols, J. D. (2004). Investigating species co‐occurrence patterns when species are detected imperfectly. Journal of Animal Ecology, 73(3), 546-555.
+
+#The 2-species occupancy model is fit in MARK. See,
+#https://sites.warnercnr.colostate.edu/gwhite/occupancy-estimation-two-species/
+
+#This model can only be fit in MARK. It can not be fit in RMark, see
+#http://www.phidot.org/forum/viewtopic.php?f=21&t=3995&p=13211&hilit=mlogit#p13211
+
+#This model is also not setup to be fit in RPresence.
+
+#Setup the workspace and load the first simulated data set
+rm(list=ls())
+load("simulation study/sim.full.data")
+length(sim.data$obs.matrix)
+q=1
+y=sim.data$obs.matrix[[1]]
+
+#Need to turn this matrix into a 2-species occupancy
+#detection matrix that is suitable for MARK
+n.occs=dim(y)[2]
+n.sites=dim(y)[1]
+night.mat=matrix(0,n.sites,n.occs)
+day.mat=matrix(0,n.sites,n.occs)
+joint=matrix(0,n.sites,n.occs*2) #This is for MARK
+for(i in 1:n.sites){
+  for(j in 1:n.occs){
+    if(y[i,j]==1){night.mat[i,j]=0;day.mat[i,j]=0}
+    if(y[i,j]==2){night.mat[i,j]=1;day.mat[i,j]=0}
+    if(y[i,j]==3){night.mat[i,j]=0;day.mat[i,j]=1}
+    if(y[i,j]==4){night.mat[i,j]=1;day.mat[i,j]=1}
+  }
+  joint[i,]=c(rbind(night.mat[i,], day.mat[i,]))  
+}
+
+night.mat
+day.mat
+joint
+
+#Create the CH inp file for MARK
+mark_input=as.data.frame(apply(joint,1,paste,collapse=""))
+mark_input <- data.frame(lapply(mark_input, as.character), stringsAsFactors=FALSE)
+
+#Output the inp file for MARK
[…]

### 2020-07-30 Model Statement/MSTOM Statement.tex
Updated MSTOM pdf

@@
 Our interest is to understand the temporal structuring of habitat use, such as what are the spatial locations (and associated probabilities) that an animal uses during the day (sunrise to sunset), during the night (sunset to sunrise), or during the day and night (sunrise to sunrise).
-The exact temporal structure ultimately depends on the diel pattern of the species; for species that are highly diurnal, temporal categories of interest could be dawn, morning, afternoon, and dusk and for species that are highly noctural, temporal categories may be dusk, early night, late night, and dusk.
+The exact temporal structure ultimately depends on the diel pattern of the species; for species that are highly diurnal, temporal categories of interest could be dawn, morning, afternoon, and dusk.
+For species that are highly nocturnal, temporal categories may be dusk, early night, late night, and dusk.
 Further, the why an animal uses a location during one time period or another is species-specific and depends on the life history strategy and behavior of the animal.
-This model could be useful to understand temporal habitat loss due to anthropogenic activity; for example, a species may use a site when considering the entire 24-hour diel period despite human influence, but change the time period of day they use the site (day only $\rightarrow$ night only) or use the site less by dropping activity during one part of the day (day and night $\rightarrow$ night only).
+This model could be useful to understand temporal habitat loss due to anthropogenic activity; for example, a species may use a site when considering the entire 24-hour diel period, but in effect change their time period of day they use the site (day only $\rightarrow$ night only) because of diurnal human activity or use the site less by dropping activity during one part of the day (day and night $\rightarrow$ night only).
 \\
@@
 The utility of this modeling framework will be well suited for remote sampling of species of interest that is continuous throughout the 24-hour diel period, such as camera traps.
-However, this type of plotless sampling in a continuous environment presents some reinterpretation of parameters as animals move in and out a camera site.
+However, this type of plotless (point-based) sampling in a continuous environment presents some reinterpretation of parameters as animals move in and out a camera site.
 Specifically, occupancy probabilities do not reflect `instantaneous occurrence', but rather `asymptotic occurrence', or the probability a site is used at least once over the duration of sampling \citep{efford2012}.
@@
 
-Below, we outline a series of multi-state temporal occupancy models (MSTOM) that can be used to understand temporal habitat-use.
-This model is a special form of the general multi-state occupancy model with state uncertainty \citep{MacKenzie2009}.
-Furthermore, the MSTOM Full model is equivalent to the original co-occurence occupancy model \citep{MacKenzie2004} with two-species and the equivalent four states.
-A demonstration of their equivalent estimation of the most probable parameters values (i.e., maximum likelihood estimates or mode of the posterior distribution) can be found in the R script, `MSTOM Full comparison with 2.species.occ.model.r'.
-Note, that while MacKenzie et al.
-2004 is known to have numerical optimization issues, especially when modeling covariates \citep{MacKenzie2004}, this issue does not persist when using a hierarchical Bayesian approach with markov chain monte carlo optimization.
+Below, we outline a series of multi-state temporal occupancy model (MSTOM) parameterizations that can be used to understand temporal habitat-use.
+The MSTOM is a special form of the general multi-state occupancy model with state uncertainty \citep{MacKenzie2009} and is equivalent to the original co-occurence occupancy model \citep{MacKenzie2004} with two-species and the equivalent four states.
+A demonstration of their equivalent estimation of the most probable parameters values (i.e., maximum likelihood estimates and mode of the posterior distributions) can be found in the R script, `MSTOM Full comparison with 2.species.occ.model.r'.
+Note, that while the likelihood models of MacKenzie et al.
+2004 are often discussed as being prone to numerical optimization issues, especially when modeling covariates \citep{MacKenzie2004}, this issue does not persist when using a hierarchical Bayesian approach with markov chain monte carlo optimization.
 
@@
 Each site is considered to be in one of $M$ mutually exclusive states during the survey period in which we can assume there are no changes in the state occurrence (i.e., closure).
-We will define four total states throughout ($M = 4$) where the states of site $i$ are 1) not occupied, 2) occupied during the day only, 3) occupied during night only, and 4) occupied during the night and day.
+We will define four total states throughout ($M = 4$) where the states of site $i$ are 1) not occupied, 2) occupied during the day only, 3) occupied during the night only, and 4) occupied during the night and day.
 State occurrence probabilities in the same order are,
@@
 Let,  $p^{m,l}_{j}$ be the probability of observing the occupancy state $l$, given the true state is $m$ in survey $j$.
-To maintain the nomenclature of our intended inference and assuming no site- or survey-level variation in detection probabilities, we can define the $p^{2,2}$ as the probability of detection during the day ($p^{\text{Day}}$), $p^{3,3}$ as the probability of detection during the night ($p^{\text{Night}}$), and the detection probability in state 4 (in order) as $pND^{1}$, $pND^{2}$, $pND^{3}$, and $pND^{4}$.
+To maintain the nomenclature of our intended inference and assuming no site- or survey-level variation in detection probabilities, we can define $p^{2,2}$ as the probability of detection during the day state ($p^{\text{Day}}$), $p^{3,3}$ as the probability of detection during the night state ($p^{\text{Night}}$), and the detection probability in state 4 (in order) as $pND^{1}$, $pND^{2}$, $pND^{3}$, and $pND^{4}$.
 The detection matrix for survey $j$ is $M$ x $M$ with the observed (columns) and true state (row) as,
@@
 
-Under this model, there is no shared relationship between the state occupancy probability 4 (night and day) and states 2 (day only) and 3 (night only).
+Under this model, there is no shared relationship between the state occupancy probability of state 4 (night and day) and states 2 (day only) and 3 (night only).
 There is also no relationship shared between the detection probabilities of the same states.
@@
 \subsubsection{Multinomial-logit Model 1}
-Instead of focusing on estimating site-level occupancy probabilities, we could recast this model using a multinomial logit link as, 
+It is common for model optimization reasons, the desire to use certain prior distributions, or when using covarites to model probabilities to recast occupancy and detection probabilities using a multinomial logit link as,
 \begin{center}
@@
 \textbf{z}_{i} &\sim \text{Categorical}(\boldsymbol{\psi})\\
-\boldsymbol{\psi}_{i} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
+\boldsymbol{\psi} &= \begin{bmatrix} \psi^1_{i} & \psi^2_{i} & \psi^3_{i} & \psi^4_{i} \end{bmatrix}.\\
 \psi^1 &=\frac{\phi^1_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}\\
@@
 \psi^4 &=\frac{\phi^4_{i}}{\phi^1_{i}+ \phi^2_{i}+\phi^3_{i}+\phi^4_{i}}.\\
-\end{align*}
-\end{center}
-
-\begin{center}
-\begin{align*}
 \phi^1 &= 1\\
@@
 
+Further, we could do the same for the detection parameters as, 
+\begin{center}
+\begin{align*}
+\boldsymbol{q} &= \begin{bmatrix} 1 & 0 & 0 & 0 \\ 
+									1 &  q^{2,2} & 0 & 0 \\ 
+									1 & 0 & q^{3,3} & 0\\
+  								       1 & q^{4,2} & q^{4,3} & q^{4,4}
+  								      \end{bmatrix}\\
+q^{2,2} &=e^{\beta^{\text{Day}}}\\
+q^{3,3} &=e^{\beta^{\text{Night}}}\\
+q^{4,2} &=e^{\beta^{\text{ND2}}}\\
+q^{4,3} &=e^{\beta^{\text{ND3}}}\\
+q^{4,4} &=e^{\beta^{\text{ND2}}+\beta^{\text{ND3}}+\beta^{\text{ND4}}}
+\end{align*}
+\end{center}
+and the probability of detection parameters can be derived as,
+
+\begin{center}
+\begin{align*}
+p^{\text{Day}} &=\frac{q^{2,2}}{1+q^{2,2}+0+0}\\
+p^{\text{Night}} &=\frac{q^{3,3}}{1+0+q^{3,3}+0}\\
+p^{\text{ND1}} &=\frac{1}{1+q^{4,2}+q^{4,3}+q^{4,4}}\\
+p^{\text{ND2}} &=\frac{q^{4,2}}{1+q^{4,2}+q^{4,3}+q^{4,4}}\\
+p^{\text{ND3}} &=\frac{q^{4,3}}{1+q^{4,2}+q^{4,3}+q^{4,4}}\\
+p^{\text{ND4}} &=\frac{q^{4,4}}{1+q^{4,2}+q^{4,3}+q^{4,4}}.
+\end{align*}
+\end{center}
 A diffuse prior for all coefficients could be done using $\text{Logistic}(0,1).$ This model is coded for JAGS in the file `jags.multistate.occ.full.alt.R'.
@@
   								      \end{bmatrix}\\
-q^{2,2}_{ij} &=e^{\alpha^{\text{Day}}_{0} +\alpha^{\text{Day}}_{1}\times x_{i}+\alpha^{\text{Day}}_{2}\times v_{ij}}\\
-q^{3,3}_{ij} &=e^{\alpha^{\text{Night}}_{0} +\alpha^{\text{Night}}_{1}\times x_{i}+\alpha^{\text{Day}}_{2}\times v_{ij}}\\
-q^{4,2}_{ij} &=e^{\alpha^{\text{ND2}}_{0} +\alpha^{\text{ND2}}_{1}\times x_{i}+\alpha^{\text{Day}}_{2}\times v_{ij}}\\
-q^{4,3}_{ij} &=e^{\alpha^{\text{ND3}}_{0} +\alpha^{\text{ND3}}_{1}\times x_{i}+\alpha^{\text{Day}}_{2}\times v_{ij}}\\
-q^{4,4}_{ij} &=e^{\alpha^{\text{ND4}}_{0} +\alpha^{\text{ND4}}_{1}\times x_{i}+\alpha^{\text{Day}}_{2}\times v_{ij}}
+q^{2,2}_{ij} &=e^{\beta^{\text{Day}}_{0} +\beta^{\text{Day}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
+q^{3,3}_{ij} &=e^{\beta^{\text{Night}}_{0} +\beta^{\text{Night}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
+q^{4,2}_{ij} &=e^{\beta^{\text{ND2}}_{0} +\beta^{\text{ND2}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
+q^{4,3}_{ij} &=e^{\beta^{\text{ND3}}_{0} +\beta^{\text{ND3}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}\\
+q^{4,4}_{ij} &=e^{\beta^{\text{ND4}}_{0} +\beta^{\text{ND4}}_{1}\times x_{i}+\beta^{\text{Day}}_{2}\times v_{ij}}
 \end{align*}
@@
 
-Diffuse priors for all $\alpha$ parameters could be specifed as Logistic(0,1).
+Diffuse priors for all $\beta$ parameters could be specified as Logistic(0,1).
 
@@
 It might be that sites used during the day and night (state 4) are not unique, but simply random and independently determined by the probability any site is used during the day ($\psi^{\text{Day.M}}$) and night ($\psi^{\text{Night.M}}$).
-Rather than exclusive state probabilities, $\psi^{\text{Day.M}}$ is the probability of using any site during the day, regardless of use or not use at night (marginal probability); further, $\psi^{\text{Night}}$ is the probability of using any site at night, regardless of use during the day.
+Rather than exclusive state probabilities, $\psi^{\text{Day.M}}$ is the probability of using any site during the day, regardless of use or not use at night (marginal probability); further, $\psi^{\text{Night.M}}$ is the probability of using any site at night, regardless of use during the day.
 We can redefine our occupancy parameters as,
@@
 \subsubsection{Multinomial-logit Mode 1}
-An alternative reduced model specification is to use the multinomial-logit parameterization, which can constrain relationships among state probabilties and include covariates.
-This model could parameterized as,
+An alternative reduced model specification is to use the multinomial-logit parameterization, which can constrain relationships among state probabilities and include covariates.
+This model could be parameterized as,
 \begin{center}
@@
 
-The detection matrix could also be specified similarly on the logit scale as,
+The detection matrix could also be specified similarly using the multinomial-logit as,
 
@@
 This model could be easily extended to include a difference of occupancy and detection in state 4 by a small modification of $\phi^4_{i} = e^{\alpha_{1}+\alpha_{2}+\alpha_{3}}$ and the matrix cell $q^{4,4}$ to $e^{\beta_{1}+\beta_{2}+\beta_{3}} $.
+This would constrain the probability of detection during the day to be the same in state 2 and state 4 and the probabilty of detection during the night to be the same in state 3 and state 4.
+But, detection of state 4 in state 4 would be allowed to be more or less than the combination of the probability of detection during the day and night alone.
 A diffuse prior for all coefficients could be done using $\text{Logistic}(0,1).$ \\

### 2020-07-30 Diel_Occ.R

@@ -1,187 +1,0 @@
-# Goals -----------------------------------------------------------------
-
-# Converting capture data (1's and 0's) into diel capture data (1,2,3,4)
-# 1 = 0 (not occupied), 2 = Only day, 3 = Only night, 4 = Day and Night use
-
-
-# soon to come ->function
-# df with columns Date, Time, Station 
-# occ_number <- function(df,lat=-15.206893, lon=49.603431, start, end, 
-#                        tz="Etc/GMT-3", occasion_step=6)
-
-# Set-up -----------------------------------------------------------------
-
-rm(list=ls())
-gc() #garbage collection
-
-library(tidyverse) # data manipulation 
-library(lubridate) # dates 
-library(suncalc) # sunlight and sunset 
-
-# R project location
-setwd("C:/Users/Kim Rivera/Documents/Summer Project/TOD data/Fosa")
-
-# load metadata 
-# Data Org = Date, Time (24 hr clock), Station (Camera) Name
-fosa=read_csv(file="AJB_2008_TOD.csv")
-fosa
-
-# Identify and delete/ edit incomplete data  
-
-# Organize Data Headers (if using read_csv)
-# colnames(fosa)=c("Date","Time","Station")
-
-
-# Organize dates ----------------------------------------------------------
-
-# ATTENTION: dates are in Eastern African Timezone 
-# if cameras malfunction and time is missing BUT
-# photos are available, use light to distinguish if photos are 'Night' or 'Day'
-# Here, raw data recorded 'Night' as PM and 'Day' as AM (see below)
-
-# mutate(date_transit = case_when(is.na(date_transit) & Time=="AM" ~ ymd_hms(Date,"12:00:00"),
-#                         is.na(date_transit) & Time=="PM" ~ ymd_hms(Date,"00:00:00"),
-#                         !is.na(date_transit) ~ date_transit))
-
-# combine date and time and set timezone
-fosa <- fosa %>% 
-  mutate(date = ymd_hms(paste(fosa$Date,fosa$Time), tz="Etc/GMT-3")) %>% 
-  arrange(date)
-
-
-# Transform time into 'Day' and 'Night' ---------------------------------------
-
[…]

### 2020-07-30 JAGS/jags.multistate.occ.full.alt.R

@@ -33,9 +33,9 @@
     q[3,j,3] <- exp(beta2)
     q[3,j,4] <- 0
     q[4,j,1] <- 1
-    q[4,j,2] <- exp(beta1)
-    q[4,j,3] <- exp(beta2)
-    q[4,j,4] <- exp(beta1 + beta2 + beta3)
+    q[4,j,2] <- exp(beta3)
+    q[4,j,3] <- exp(beta4)
+    q[4,j,4] <- exp(beta3 + beta4 + beta5)
     }
     # State-space likelihood
     # State equation: model of true states (z)

### 2020-07-30 Data Processing/Diel_Occ_Func.R
Updated Kim's data processing script
Kim's data processing script

@@ -22,114 +22,29 @@ library(suncalc) # sunlight and sunset
 # Set your variables -----------------------------------------------------------
 time.zone = "Etc/GMT-3" # Time zone of your study site
 latitude = -15.206893 # Latitude of study site
-longitude = 49.603431 # Longitude of sudy site
-start.survey = "2008-09-02" # Starting date of survey
-end.survey = "2008-11-13" #Ending date of survey
+longitude = 49.603431 # Longitude of study site
+start.survey = "2008-09-02" # Starting date of survey (YYYY-MM-DD)
+end.survey = "2008-11-13" #Ending date of survey (YYYY-MM-DD)
 occasion.length = 6 # Number of days per occasion 
-station.ID = "AJB" # Name of Station 
-photo.data=read_csv(file="C:/Users/Kim Rivera/Documents/Summer Project/TOD data/AJB/Fosa/AJB_2008_TOD.csv") # CSV of 'Date', 'Time', and 'Station', of detections
-detection.matrix=read_csv(file="C:/Users/Kim Rivera/Documents/Summer Project/Capture Data/AJB/2008_occ.csv", na = ".") # Detection matrix
+station.ID = "AJB" # Name of Study Area
+detection.matrix=read_csv(file="Data Processing/Example_Detection_Matrix.csv") # Original Detection matrix
+photo.data=read_csv(file="Data Processing/Example_Photo_Data.csv", na = ".") # CSV of 'Date', 'Time', and 'Station', of defections
 
+#Note- the detection matrix can be 1 (detected), 0 (not detected), and . (not surveyed).
+#      The first row of the detection matrix and photo.data are column names       
+
+#Note- the photo.date dates need to be as YYYY-MM-DD
 # Run the function -------------------------------------------------------------
-diel_occ = function(photo.data, 
-                    detection.matrix, 
-                    time.zone, 
-                    latitude, 
-                    longitude, 
-                    start.survey, 
-                    end.survey, 
-                    station.ID,
-                    occasion.length){
-  # Organize dates -------------------------------------------------------------
-  # Here we combine and format the date-time in our defined time zone
-  photo.data <- photo.data %>% 
-    mutate(date = ymd_hms(paste(photo.data$Date,photo.data$Time), tz=time.zone)) %>% 
-    arrange(date)
-  
-  # Transform time into 'Day' and 'Night' --------------------------------------
[…]

### 2020-07-30 Data Processing/diel.occ.fun.R

@@ -1,0 +1,98 @@
+diel.occ.fun = function(photo.data, 
+                    detection.matrix, 
+                    time.zone, 
+                    latitude, 
+                    longitude, 
+                    start.survey, 
+                    end.survey, 
+                    station.ID,
+                    occasion.length){
+  # Organize dates -------------------------------------------------------------
+  # Here we combine and format the date-time in our defined time zone
+  photo.data <- photo.data %>% 
+    mutate(date = ymd_hms(paste(photo.data$Date,photo.data$Time), tz=time.zone)) %>% 
+    arrange(date)
+  
+  # Transform time into 'Day' and 'Night' --------------------------------------
+  # This function gives us the sunrise and sunset based on our study site's lat/long 
+  sun_t = getSunlightTimes(date =(photo.data$Date), lat = latitude, 
+                           lon = longitude, keep = c("sunrise", "sunset"),
+                           tz=time.zone) %>% 
+    select(sunrise, sunset)
+  
+  # Add sunrise and sunset to the metadata 
+  photo.data <- tibble(cbind(photo.data, sun_t))
+  
+  # Create a new column, 'type,' which classifies time as 'Day' and'Night'
+  photo.data <- photo.data %>%
+    mutate(type=ifelse(date > sunrise & date < sunset, "Day", "Night"))
+  
+  # Add occasion number --------------------------------------------------------
+  # create a vector with all the dates 
+  start.survey = ymd(start.survey)
+  end.survey = ymd(end.survey)
+  
+  full_date <- seq(start.survey, end.survey, by="day") 
+  occ_length <- ceiling(length(full_date)/occasion.length) 
+  occ <- sort(rep(1:occ_length,length.out=length(full_date))) # index of occasions
+  
+  full_date <- tibble(Date = full_date, occ) # combine days and occasion numbers
+  photo.data <- left_join(photo.data, full_date) # add the occasion to the metadata
+  
+  # Build new diel detection matrix ----------------------------------------------------
+  # add a column for stations (1ABJ01 to 1ABJ..)
+  # add 0's to numbers from 1 to 9
+  station.num <- 1:nrow(detection.matrix)
+  if (length(station.num) <= 10){
+    station.num <- paste("0",station.num)
+  } else {
[…]

### 2020-07-30 Data Processing/Diel.Occ.Script.R
name change

@@ -1,0 +1,50 @@
+# Function Purpose--------------------------------------------------------------
+# The diel_occ function allows single-state occupancy detection matrices to 
+# be converted into a 4 state matrix in order to better understand wildlife
+# diel activity. Rather than 0's indicating no detection and 1's indicating
+# an animal was detected, diel states are defined as the following:
+# 1 = not detected
+# 2 = detected during the day only (after sunrise, before sunset)
+# 3 = detected during the night only (after sunset, before sunrise)
+# 4 = detected during the day AND night 
+
+# Lets get started!
+
+# Clean Up ---------------------------------------------------------------------
+rm(list=ls())
+gc() #garbage collection
+
+# Required Libraries -----------------------------------------------------------
+library(tidyverse) # data manipulation 
+library(lubridate) # dates 
+library(suncalc) # sunlight and sunset 
+
+# Set your variables -----------------------------------------------------------
+time.zone = "Etc/GMT-3" # Time zone of your study site
+latitude = -15.206893 # Latitude of study site
+longitude = 49.603431 # Longitude of study site
+start.survey = "2008-09-02" # Starting date of survey (YYYY-MM-DD)
+end.survey = "2008-11-13" #Ending date of survey (YYYY-MM-DD)
+occasion.length = 6 # Number of days per occasion 
+station.ID = "AJB" # Name of Study Area
+detection.matrix=read_csv(file="Data Processing/Example_Detection_Matrix.csv") # Original Detection matrix
+photo.data=read_csv(file="Data Processing/Example_Photo_Data.csv", na = ".") # CSV of 'Date', 'Time', and 'Station', of defections
+
+#Note- the detection matrix can be 1 (detected), 0 (not detected), and . (not surveyed).
+#      The first row of the detection matrix and photo.data are column names       
+
+#Note- the photo.date dates need to be as YYYY-MM-DD
+# Run the function -------------------------------------------------------------
+source("Data Processing/diel.occ.fun.R")
+
+new.det.matrix=diel.occ.fun(photo.data,
+                        detection.matrix, 
+                        time.zone, 
+                        latitude, 
+                        longitude, 
+                        start.survey, 
+                        end.survey, 
+                        station.ID, 
+                        occasion.length)
+ 
+new.det.matrix

### 2020-07-30 Data Processing/Diel_Occ_Func.R

@@ -1,50 +1,0 @@
-# Function Purpose--------------------------------------------------------------
-# The diel_occ function allows single-state occupancy detection matrices to 
-# be converted into a 4 state matrix in order to better understand wildlife
-# diel activity. Rather than 0's indicating no detection and 1's indicating
-# an animal was detected, diel states are defined as the following:
-# 1 = not detected
-# 2 = detected during the day only (after sunrise, before sunset)
-# 3 = detected during the night only (after sunset, before sunrise)
-# 4 = detected during the day AND night 
-
-# Lets get started!
-
-# Clean Up ---------------------------------------------------------------------
-rm(list=ls())
-gc() #garbage collection
-
-# Required Libraries -----------------------------------------------------------
-library(tidyverse) # data manipulation 
-library(lubridate) # dates 
-library(suncalc) # sunlight and sunset 
-
-# Set your variables -----------------------------------------------------------
-time.zone = "Etc/GMT-3" # Time zone of your study site
-latitude = -15.206893 # Latitude of study site
-longitude = 49.603431 # Longitude of study site
-start.survey = "2008-09-02" # Starting date of survey (YYYY-MM-DD)
-end.survey = "2008-11-13" #Ending date of survey (YYYY-MM-DD)
-occasion.length = 6 # Number of days per occasion 
-station.ID = "AJB" # Name of Study Area
-detection.matrix=read_csv(file="Data Processing/Example_Detection_Matrix.csv") # Original Detection matrix
-photo.data=read_csv(file="Data Processing/Example_Photo_Data.csv", na = ".") # CSV of 'Date', 'Time', and 'Station', of defections
-
-#Note- the detection matrix can be 1 (detected), 0 (not detected), and . (not surveyed).
-#      The first row of the detection matrix and photo.data are column names       
-
-#Note- the photo.date dates need to be as YYYY-MM-DD
-# Run the function -------------------------------------------------------------
-source("Data Processing/diel.occ.fun.R")
-
-new.det.matrix=diel.occ.fun(photo.data,
-                        detection.matrix, 
-                        time.zone, 
-                        latitude, 
-                        longitude, 
-                        start.survey, 
-                        end.survey, 
-                        station.ID, 
-                        occasion.length)
- 
-new.det.matrix

### 2020-07-30 Data Processing/Fosa_Occupancy.R
Create Fosa_Occupancy.R

@@ -1,0 +1,419 @@
+rm(list=ls())
+setwd("C:/Users/Kim Rivera/Documents/Summer Project/Capture Data")
+
+#############################################################################################
+################################ 2008 #######################################################
+
+fosa_2008=read.csv(file="2008_occ.csv", head=TRUE)
+head(fosa_2008)
+
+colnames(fosa_2008)<-paste("Occ", 1:ncol(fosa_2008), sep=" ")
+rownames(fosa_2008)<-paste("1AJB", 1:nrow(fosa_2008), sep=" ")
+
+library(dplyr)
+fosa_2008 = na_if(fosa_2008[,], ".")
+
+fosa_mat_2008=as.matrix(fosa_2008)
+fosa_mat_2008
+
+## Site Temporal Sub ##
+
+#  1 = 0 (not occupied), 2 = Only day, 3 = Only night, 4 = Day and Night use
+# first chage 1's to 2/3/4, then change all remaining 0's to 1's
+
+fosa_mat_2008[12,1] = gsub("1", "3", fosa_mat_2008[12,1]) # Occ 1
+fosa_mat_2008[12,2] = gsub("1", "3", fosa_mat_2008[12,2]) # Occ 2
+fosa_mat_2008[12,4] = gsub("1", "3", fosa_mat_2008[12,4]) # Occ 4
+fosa_mat_2008[12,5] = gsub("1", "2", fosa_mat_2008[12,5]) # Occ 5
+fosa_mat_2008[12,6] = gsub("1", "4", fosa_mat_2008[12,6]) # Occ 6
+fosa_mat_2008[12,7] = gsub("1", "3", fosa_mat_2008[12,7]) # Occ 7
+fosa_mat_2008[12,12] = gsub("1", "3", fosa_mat_2008[12,12]) # Occ 8
+
+fosa_mat_2008[1,6] = gsub("1", "3", fosa_mat_2008[1,6]) # Occ 6
+
+fosa_mat_2008[2,11] = gsub("1", "3", fosa_mat_2008[2,11]) # Occ 11
+
+fosa_mat_2008[3,8] = gsub("1", "3", fosa_mat_2008[3,8]) # Occ 8
+
+fosa_mat_2008[8,4] = gsub("1", "4", fosa_mat_2008[8,4]) # Occ 4
[…]

### 2020-07-30 simulation study/sim.data.MSTOM.full.model.r
error in det.matrix.func

@@ -83,7 +83,7 @@ pND.0=1-pND.ND- pND.D-pND.N                   #Detection of state 1 in state 4
 
 #This function creates the detection matrix
 source("simulation study/det.matrix.func.r")
-det.matrix=det.matrix.func(pNight,pDay,pND.N,pND.D,pND.0)
+det.matrix=det.matrix.func(pNight,pDay,pND.N,pND.D,pND.ND,pND.0)
 det.matrix
 
 #Check to make sure the sums of the rows (true states) equal one and no negative values

### 2020-07-30 simulation study/sim.data.MSTOM.null.model.r

@@ -94,7 +94,7 @@ pND.N=omega.det[3]
 
 #This function creates the detection matrix
 source("simulation study/det.matrix.func.r")
-det.matrix=det.matrix.func(pNight,pDay,pND.N,pND.D,pND.0)
+det.matrix=det.matrix.func(pNight,pDay,pND.N,pND.D,pND.ND,pND.0)
 det.matrix
 
 #Check to make sure the sums of the rows (true states) equal one and no negative values

### 2020-11-13 RNP Fosa/RNP.MSTOM.modeling.script.r
Add RNP script
add RNP script and update Model Statement

@@ -30,9 +30,16 @@ source("RNP Fosa/CPO.function.r")
 load("RNP Fosa/RNP2.data")
 
 #assign data to objects
-y=RNP.data[[1]] #detection history
-covs=RNP.data[[2]]
+y=RNP2.data[[1]] #detection history
+covs=RNP2.data[[2]]
 
+#Derive Site/Survey Covariate
+VOH=1:26
+VA=27:53
+CVB=54:95
+survey=c(rep("VOH",length(VOH)),rep("VA",length(VA)),rep("CVB",length(CVB)))
+X.survey=model.matrix(~survey,contrasts = list(survey = "contr.sum"))
+head(X.survey)
 #################################################################
 #################################################################
 

### 2020-11-13 RNP Fosa/RNP.MSTOM.modeling.script2.r

@@ -1,0 +1,587 @@
+#################################################################
+#################################################################
+#This script fits the Ranomafana National Park fosa occupancy camera trap data - Data from Gerber et al. 2012
+#Gerber, B. D., Karpanty, S. M., & Randrianantenaina, J. (2012). The impact of forest logging and fragmentation on carnivore species composition, density and occupancy in Madagascar's rainforests. Oryx, 46(3), 414-422.
+#
+#Specifically, it uses the camera surveys at Sahamalaotra and Valohoaka-Vatoharanana within the park.
+#The states are assigned as,
+
+# 4 States:
+# 1: No use
+# 2: Day only
+# 3: Night Only
+# 4: Night and Day
+
+
+#Day was assigned as the period between sunrise and sunset
+#Nigt was assigned as the period between sunset and sunrise
+
+#*****THIS script is only different from the first modeling script because
+#it uses a categorical effect for survey in the same models
+
+rm(list=ls())
+logit=function(x){log(x/(1-x))}
+expit=function(x){exp(x)/(exp(x)+1)}
+library(rjags)
+library(runjags)
+library(coda)
+source("RNP Fosa/multi.state.likelihood.r")
+source("RNP Fosa/CPO.function.r")
+
+#load the prepared data file
+#load("RNP Fosa/RNP.data")
+load("RNP Fosa/RNP2.data")
+
+#assign data to objects
+y=RNP2.data[[1]] #detection history
+covs=RNP2.data[[2]]
+
+#Derive Site/Survey Covariate
+VOH=1:26
+VA=27:53
+CVB=54:95
+#survey=c(rep("VOH",length(VOH)),rep("VA",length(VA)),rep("CVB",length(CVB)))
+survey=c(rep("2007",length(VOH)+length(VA)),rep("2008",length(CVB)))
+X.survey=model.matrix(~survey,contrasts = list(survey = "contr.sum"))
+head(X.survey)
+#################################################################
+#################################################################
+
+#Examine covariate data
+
+#FYI-The Locals covariate- all detections of people are during the day. There are no nighttime detections.
+#Locals covariate is probably not a good measure of overall human/forest disturbance.
+
+#Look at covariate correlations
+cor(covs$DistMatrix,covs$Locals)
[…]

### 2020-11-13 RNP Fosa/RNP.MSTOM.plotting.r

@@ -9,11 +9,11 @@ library(ggplot2)
 
 
 #for first fit
-#load("RNP Fosa/M1.fit")
+load("RNP Fosa/M1.fit")
 
-rm(list=ls())
-load("RNP Fosa/M1.full.out")
-fit <- combine.mcmc(M1.full)
+#rm(list=ls())
+#load("RNP Fosa/M3.red.out")
+#fit <- combine.mcmc(M3.red)
 #load the prepared data file
 load("RNP Fosa/RNP2.data")
 covs=RNP2.data[[2]]

### 2020-11-13 RNP Fosa/RNP.MSTOM.plotting.temp.r

@@ -1,0 +1,166 @@
+#M1 is the most supported model
+#Load the model and plots results
+
+library(rjags)
+library(runjags)
+library(coda)
+library(bayesplot)
+library(ggplot2)
+
+
+#for first fit
+load("RNP Fosa/RNP Fit Original/M3.red.out")
+
+#rm(list=ls())
+#load("RNP Fosa/M3.red.out")
+#fit <- combine.mcmc(M3.red)
+#load the prepared data file
+load("RNP Fosa/RNP2.data")
+covs=RNP2.data[[2]]
+cov=covs$DistTown
+hist(cov)
+cov1.unscaled=as.numeric(cov)
+cov1.scaled=as.numeric(scale(cov,center = TRUE,scale = TRUE))
+
+
+#Plot detection parameters
+pDay=fit[,which(grepl("pDay",colnames(fit)))]
+pNight=fit[,which(grepl("pNight",colnames(fit)))]
+
+pND.0=fit[,which(grepl("pND",colnames(fit)))[1]]
+pND.D=fit[,which(grepl("pND",colnames(fit)))[2]]
+pND.N=fit[,which(grepl("pND",colnames(fit)))[3]]
+pND.ND=fit[,which(grepl("pND",colnames(fit)))[4]]
+
+det.matrix=cbind(pDay,
+                 pNight,
+                 pND.0,
+                 pND.D,
+                 pND.N,
+                 pND.ND)
+
+par(mfrow=c(1,2))
+plot(density(pDay),xlim=c(0,0.2),lwd=3)
+lines(density(pNight),xlim=c(0,0.2),lwd=3,col=2)
+plot(density(pND.D),xlim=c(0,0.2),lwd=3)
+lines(density(pND.N),xlim=c(0,0.2),lwd=3,col=2)
+lines(density(pND.ND),xlim=c(0,0.2),lwd=3,col=3)
+
+
+png(file="RNP Fosa/RNP.fosa.det.parms.png",res=200,units = "in",height=8,width=8)
+color_scheme_set("red")
+mcmc_intervals(det.matrix, pars = colnames(det.matrix)[-3],xlab="Detection Parameters")+ labs(x = "Detection Probability",y="State")+
+     theme(text = element_text(size=20))
+dev.off()
+
+colnames(fit)
+#The first 6 columns are the alpha parameters
+fit.matrix=as.matrix(fit)
+
[…]
